### 2023-03-30 paper/mabs2023.tex
Initial commit

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{tabularx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\begin{document}
%
\title{Towards automated provenance collection for experimental runs of agent-based models\thanks{Supported by The Scottish Government}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Doug Salt\inst{1}\orcidID{0000-1111-2222-3333} \and
Gary Polhill\inst{2,3}\orcidID{1111-2222-3333-4444} }
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{The James Hutton Institute, Craigiebuckler Aberdeen AB15 8QH Scotland
\email{doug.salt@hutton.ac.uk}\\
\url{http://hutton.ac.uk}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
The abstract should briefly summarize the contents of the paper in
15--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}

Replication of social simulation results has been highlighted as a significant
issue for the ABM for a number of years (e.g.  \cite{edmonds2003replication}),
and in particular the paper that forms the basis of this work
\cite{polhill2017lessons}.  The focus of replication work has previous just
been on the model itself, but as was shown in ibid, the analysis of the
outputs of the model can potentially be just as complex, and no less difficult
to replicate unless adequate records are kept. Indeed there as an additional
attempt to reproduce the results and even with the lessons learnt from ibid, it
still took months to recreate the final diagrams that were submitted to paper.
The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
some guidance highlighting the need to keep a notebook of the analysis done and
a standardised approach to making that notebook; however, the lessons learned
from the replication exercise in \cite{polhill2017lessons} show that more
detailed guidance on the information that should be recorded.  and on tools
that could be used to support the process. Ideally the aim should be to
completely automate the process, record accurately sources of data used,
version check applications used essentially providing a complete graph from
data to result.

The output analysis replication in this paper concerns earlier work with
FEARLUS-SPOMM, which is a coupled agent-based model of agricultural decision-
making and species stochastic patch occupancy metacommunity model that has been
used to explore incentivisation strategies to improve biodiversity (Polhill et
al. 2013; Gimona and Polhill 2011). Belonging to the ‘typification’ class of
social simulations adfa, p. 1, 2011.  © Springer-Verlag Berlin Heidelberg 2011
(“theoretical constructs intended to investigate some properties that apply to
a wide range of empirical phenomena that share some common features” – Boero
and Squaz- zoni, 2005). This work involved the analysis of the outputs from
around 20,000 runs of the model using a number of techniques aimed at
demonstrating nonlinearities in the relationship between incentivisation and
biodiversity outcome.  Recording workflow data on the process used to create
analysis can be challenging, and currently there are no codified standards as
to how this should be done for ABMs. For FEARLUS-SPOMM, the methods used drew
heavily on statistical techniques available as R packages that are as part of
core R functionality. Although R allows transcripts of interactive terminal
sessions to be saved, the work involved great deal of exploration of different
ways of attempting to visualise and analyse the nonlineari- ties in the model
results, not all of which were likely to be reported in the paper. Such logs
are therefore not the best way to record the means by which the outputs were
analysed, and hence the strategy used was to save each analysis or
visualisation in a(n R) script. Since the output from the (Swarm) software that
generated the output data being analysed used a mixture of text formats, some
Perl scripts were also written to process that output into a CSV file for easy
import into R. When the MIRACLE pro- ject (Parker et al. 2015) provided a
context in which the replication of that analysis was necessary, an opportunity
was created to test the viability of the above strategy.

In the rest of this paper, we describe  a tool for automatically recording
metadata, which can be incorporated into the analysis replication process, and
how this  was used to regenerate some of the figures in Polhill et al. (2013)
and we test a tool to support keeping the records needed to per- form
replication more easily, before making some recommendations in concluding
remarks.

\section{Method}

ne of the aims was to leave as much of the code as unchanged as possible (this differs in approach to FAIR-CLI. develop this, both here and in the discussion.

One of the artefacts of the MIRACLE project \cite{} was the Social Simulation REplication Interface or SSREPI. This is the schema shown 
in \ref{fig:schema}. This schema has evolved since it original conception.
The newest version of this document may be found here \cite{}. This is a
database schema derived from the Dublin Core \cite{weibel2000dublin}, the
standard XML datatypes \cite{biron2004xml} and the PROV-O ontologies \cite{missier2013w3c}. The schema is designed with
agnosticisim towards the underlying database technology and as such has
been implmented both in PostgresSQL \cite{stonebraker1991postgres} and Sqlite3 \cite{sqliteorg2023syntax}.

\begin{figure}
\includegraphics[width=\textwidth]{img/schema.jpeg}
\caption{SSREPI Schema} \label{fig:schema}
\end{figure}

There are two important dimensions of distinction to this schema. First is
fine- versus coarse- grained metadata. Second is provenance versus
workflow. Coarse-grained metadata describes how particular files come (or
came) into being, or were (or could be) used to bring other files into
being. Fine-grained metadata describes specific values recorded in social
simulation outputs. To make the distinction concrete, suppose a simulation
produces a CSV file. The data within the CSV file are covered by
fine-grained metadata, whilst the fact that the simulation produces the
CSV file is coarse-grained. Turning to the other dimension, provenance
metadata describes what actually happens (run W of simulation X produced
output file Y), whilst workflow metadata describes what could happen
(simulation X produces an output file of type Z). The distinctions are
summarised in \ref{fig:finegrain} .

\begin{figure}
\includegraphics[width=\textwidth]{img/fine-grain-vs-coarse-grain.jpeg}
\caption{Fine grain vs coarse grain provenance} \label{fig:finegrain}
\end{figure}

For the purposes of this short paper we are concentrating on a
demonstration of recording provenance. Indeed we use the example,
mentioned in the introduction of \cite{}, and modified the original Bash \cite{} scripts to include the what we denote as \textit{primitives} . So an example of the code

In order to record provenance we 
We coded the following primitives from
the SSREPI interface definition \cite{}. These are implemented in a mixture of Python \cite{} and Bash \cite{}

\begin{table}
\caption{Table captions should be placed above the
tables.}\label{tab1}
    \begin{tabularx}{\textwidth}{|l|p{2cm}|p{6cm}|}
\hline
Primitive & Type & Purpose \\
\hline
        SSREPI\_require\_minimum & Metadata & Lower bound on software hardware required \\ SSREPI\_require\_exact & Metadata & Exact bound on software hardware required \\
        SSREPI\_application & Provenance \& Metadata & specifies some executable \\
        SSREPI\_me & Provenance \& Metadata & Determines executable being run or returns a proper reference to the executable being run. \\
        SSREPI\_run & Provenance \& metadata & Blocking invocation of an executable which will allow the specification of all inputs, outputs and arguments. Creates run-time provenance information as well \\
        SSREPI\_batch & Provenance \& metadata & Non blocking invocation of an executable which will allow the specification of all inputs, outputs and arguments. Creates run-time provenance information as well \\
        SSREPI\_argument & Provenance & An argument type to an exectuable \\
        SSREPI\_output & Provenance & An output type from an executable \\
        SSREPI\_input & Provenance & An input type for an exectuable \\
        SSREPI\_hutton\_person & Metadata & Uses our institutions databases to populate metadata for a given individual \\
        SSREPI\_person & Metdata & Provide metadata for a particular actor within this system\\
        SSREPI\_project & Metadata & Specifies a project which contains all studies \\
        SSREPI\_study & Metadata & A set of experiments makes up a single study \\
        SSREPI\_set & Metdata & Sets the default licence and other metadata \\
        SSREPI\_involvement & Metadata & Links personnel to a study \\
        SSREPI\_paper & Metadata & A paper associated with this study \\
        SSREPI\_make\_tag & Metadata & Used for building a folksonomy \\
        SSREPI\_tag & Metadata & Used to tag any entity with a folksonomy tag \\
        SSREPI\_contributor & Metadata & A  person with some kind of relation to an executable or script. \\

        SSREPI\_statistical\_method & Metadata & A statistical method is an approach to computing some statistics. It may be implemented in or as part of an application. A statistical method generates one or more statistical variables as its results, and may use the results of another statistical method in its computation. For example, computing the standard deviation of some data uses the mean of those data. \\
        SSREPI\_visualisation & Metadata & A visualisation is the process of creating an image to depict one or more (typically more than one) values. The results of a visualisation appear in a container.\\
        SSREPI\_statistics & Metadata & Statistics are activities that compute  and populate the values of statistical rvariables. They operate on raw data that are retrieved from the values using a query. To replicate a set of statistics, the query can be rerun, selecting values that are pointed to by containers entries \\
        SSREPI\_visualisation\_method & Metadata & This describes methods for generating visualisations, which then may appear in the content of a container produced by a process running an application that implements it. \\

        SSREPI\_implements & Metadata & Links a statistical or visualisation method to an application \\
        SSREPI\_parameter & Metadata & A Parameter is the name of a parameter taken by a statistical or visualisation method, used to configure the way it behaves. \\
        SSREPI\_statistical\_variable & Metadata &  A name for (one of) the result(s) of a statistical method. \\
        SSREPI\_visualisation\_variable & Provenance \& Metadata & Declares a named variable of interest \\
        SSREPI\_variable & Metadata & Names a variable of interest \\
        SSREPI\_statistical\_variable\_value & Provenance \& Metdata & Sets an actual value for a named statistical variable \\
        SSREPI\_value & Provenance & Sets an actual value. This can be connected to any metadata value such         \\ 
        SSREPI\_content & Metadata & Links a kind of output/input/argument to a variable  \\
        SSREPI\_person\_makes\_assumption & Metadata & links a person to an assumption \\
\hline
\end{tabularx}
\end{table}

Broadly speaking SSREPI\_application, SSREPI\_run, SSREPI\_batch, SSREPI\_input,
SSREPI\_output and SSREPI\_argument are responsible for recording coarse grain
provenance. SSREPI\_value, SSREPI\_visualisation\_variable\_value, SSREPI\_statistical\_variable\_value, SSREPI\_run and SSREPI\_batch record fine-grain provenance. The remaining
primitives are largely about recording metadata.

Elaboration of all this may be found in the man pages and design documents in the public repository at git@github.com:DougSalt/ABM-metadata.git

\begin{figure}
\includegraphics[width=\textwidth]{img/workflow.pdf}
\caption{The workflow sub-graph} \label{fig:workflow.pdf}
\end{figure}

\section{Results}

Obviously with complicated diagrams result from the 20,000 runs. The challenge becomes how to visualise these.
There are several graphs produced so far, these being

We show the workflow graph in fig. \ref{fig:workflow} and a section of the
proveance graph in fig. \ref{fig:sub-provenance}

\begin{figure}
\includegraphics[width=\textwidth]{img/subsection-of-provenance.png}
\caption{A small sub-section of the proevenance graph} \label{fig:sub-provenance}
\end{figure}

\section{Discussion}

What we have learnt from this 
This is a lot of pfaff. and is still not oven ready

So what use is this provenence meta data. Tracing bad data.

We demonstrate this by tracing a series of bad data across the diagrams.

The proveance should be a directed acyclic graph. There may well be loops in the schema as it stands. The authors have done their best to normalise it to uncover such cycles, but in a schema this complex they may well still exist.

Eventually we should like to take several databases run some machine learning across them to see if there are any commonalities in experiment set up and post processing.

Additionally we want to produce modules in other languages as well, such as R, python and Julia. Especially the latter.

One of the nice things was that I was able to parallelise the post processing job simply.

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography[keyword={github.com/DougSalt/MABS2023}]{global.bib}
%
\end{document}

### 2023-04-11 paper/paper.tex
Got the references correct. I need another 4 pages
content and some pretty diagrams.

--- 2023-03-30 paper/mabs2023.tex
@@
 and in particular the paper that forms the basis of this work
-\cite{polhill2017lessons}.  The focus of replication work has previous just
+\cite{polhill2017lessons}.  The focus of replication work has previously just
 been on the model itself, but as was shown in ibid, the analysis of the
@@
 from the replication exercise in \cite{polhill2017lessons} show that more
-detailed guidance on the information that should be recorded.  and on tools
+detailed guidance on the information that should be recorded,  and on tools
 that could be used to support the process. Ideally the aim should be to
@@
 making and species stochastic patch occupancy metacommunity model that has been
-used to explore incentivisation strategies to improve biodiversity (Polhill et
-al. 2013; Gimona and Polhill 2011). Belonging to the ‘typification’ class of
-social simulations adfa, p. 1, 2011.  © Springer-Verlag Berlin Heidelberg 2011
-(“theoretical constructs intended to investigate some properties that apply to
-a wide range of empirical phenomena that share some common features” – Boero
-and Squaz- zoni, 2005). This work involved the analysis of the outputs from
-around 20,000 runs of the model using a number of techniques aimed at
-demonstrating nonlinearities in the relationship between incentivisation and
-biodiversity outcome.  Recording workflow data on the process used to create
-analysis can be challenging, and currently there are no codified standards as
-to how this should be done for ABMs. For FEARLUS-SPOMM, the methods used drew
-heavily on statistical techniques available as R packages that are as part of
-core R functionality. Although R allows transcripts of interactive terminal
-sessions to be saved, the work involved great deal of exploration of different
-ways of attempting to visualise and analyse the nonlineari- ties in the model
-results, not all of which were likely to be reported in the paper. Such logs
-are therefore not the best way to record the means by which the outputs were
-analysed, and hence the strategy used was to save each analysis or
-visualisation in a(n R) script. Since the output from the (Swarm) software that
-generated the output data being analysed used a mixture of text formats, some
-Perl scripts were also written to process that output into a CSV file for easy
-import into R. When the MIRACLE pro- ject (Parker et al. 2015) provided a
-context in which the replication of that analysis was necessary, an opportunity
-was created to test the viability of the above strategy.
+used to explore incentivisation strategies to improve biodiversity
+\cite{polhill_nonlinearities_2013}, \cite{gimona2011exploring}. Belonging to the
+‘typification’ class of social simulations  \cite{boero2005does}.  This work
+involved the analysis of the outputs of approximately 20,000 runs of the model
+using a number of techniques aimed at demonstrating nonlinearities in the
+relationship between incentivisation and biodiversity outcome.  Recording
+workflow data on the process used to create analysis can be challenging, and
+currently there are no codified standards as to how this should be done for
+ABMs. For FEARLUS-SPOMM, the methods used drew heavily on statistical
+techniques available as R packages that are as part of core R functionality.
+Although R allows transcripts of interactive terminal sessions to be saved, the
+work involved great deal of exploration of different ways of attempting to
+visualise and analyse the nonlinearities in the model results, not all of
+which were likely to be reported in the paper. Such logs are therefore not the
+best way to record the means by which the outputs were analysed, and hence the
+strategy used was to save each analysis or visualisation in a(n R) script.
+Since the output from the (Swarm) software that generated the output data being
+analysed used a mixture of text formats, some Perl scripts were also written to
+process that output into a CSV file for easy import into R. When the MIRACLE pr
+ject \cite{parker2019final} provided a context in which the replication of that
+analysis was necessary, an opportunity was created to test the viability of the
+above strategy.
 
@@
 metadata, which can be incorporated into the analysis replication process, and
-how this  was used to regenerate some of the figures in Polhill et al. (2013)
-and we test a tool to support keeping the records needed to per- form
-replication more easily, before making some recommendations in concluding
-remarks.
+how this  was used to regenerate some of the figures in
+\cite{polhill_nonlinearities_2013} and we test a tool to support keeping the
+records needed to per- form replication more easily, before making some
+recommendations in concluding remarks.
 
@@
 
-ne of the aims was to leave as much of the code as unchanged as possible (this differs in approach to FAIR-CLI. develop this, both here and in the discussion.
+One of the artefacts of the MIRACLE project \cite{parker2019final} was the
+Social Simulation REplication Interface or SSREPI. This is the schema shown in
+\ref{fig:schema}.  This schema has evolved since it original conception.  The
+newest version of this document may be found here \cite{polhill2022miracle}. This is a database
+schema derived from the Dublin Core \cite{weibel2000dublin}, the standard XML
+datatypes \cite{biron2004xml} and the PROV-O ontologies \cite{missier2013w3c}.
+The schema is designed with agnosticisim towards the underlying database
+technology and as such has been implmented both in PostgresSQL
+\cite{stonebraker1991postgres} and Sqlite3 \cite{sqliteorg2023syntax}.
 
-One of the artefacts of the MIRACLE project \cite{} was the Social Simulation REplication Interface or SSREPI. This is the schema shown 
-in \ref{fig:schema}. This schema has evolved since it original conception.
-The newest version of this document may be found here \cite{}. This is a
-database schema derived from the Dublin Core \cite{weibel2000dublin}, the
-standard XML datatypes \cite{biron2004xml} and the PROV-O ontologies \cite{missier2013w3c}. The schema is designed with
-agnosticisim towards the underlying database technology and as such has
-been implmented both in PostgresSQL \cite{stonebraker1991postgres} and Sqlite3 \cite{sqliteorg2023syntax}.
+One of the aims was to leave as much of the code, i.e. the original experiment
+as unchanged as possible. That is we wanted to change the framework code rather
+than the core code. The core code in the example above being the executable,
+precompiled in Objective-C \cite{objectivec}. Additionally the support code, i.e. the
+code that prepared the data, a Perl \cite{perl563} script and the code that did the
+post analysis, a mixture of Perl and R \cite{R422}.  This meant we developed the
+database access in Python \cite{python368}, wrote wrapper scripts in Bash \cite{bash4420}. This
+differs in approach to FAIR data piperline \cite{mitchell2022fair} in which the provenance framework is
+embedded in the code that constitutes the body of the experiment as it were.
+
+With the tool in the current state this involved labouriously going through the
+code and inserting the some of the following Bash-based commands, shown in
+table \ref{tab:commands}. These commands have a fairly complex parameter set as
+it stands, and no way of checking whether the code worked other than running
+it, and this led to one of the immediate problems with this method.
 
@@
 demonstration of recording provenance. Indeed we use the example,
-mentioned in the introduction of \cite{}, and modified the original Bash \cite{} scripts to include the what we denote as \textit{primitives} . So an example of the code
+mentioned in the introduction of \cite{polhill2022miracle}, and modified the original Bash \cite{bash4420} scripts to include the what we denote as \textit{primitives} . So an example of the code
 
@@
 We coded the following primitives from
-the SSREPI interface definition \cite{}. These are implemented in a mixture of Python \cite{} and Bash \cite{}
+the SSREPI interface definition \cite{polhill2022miracle}. These are implemented in a mixture of Python and Bash.
 
@@
 \caption{Table captions should be placed above the
-tables.}\label{tab1}
+tables.}\label{tab:commands}
     \begin{tabularx}{\textwidth}{|l|p{2cm}|p{6cm}|}
@@
 
+As mentioned the method undertaken in this paper involved inserting the correct
+Bash-based commands, listed in table \ref{tab:commands}. These commands
+have a fairly complex parameter set as it stands, and no way of checking
+whether the code worked other than running it, and this led to one of the
+immediate problems with this method. The only way to see if the code was
+working was to run it, a suck it and see methodology. This was extremely time consuming. We took the obvious approach and ran small subsets of
+the ~20,000 runs or so, to see if we could get the framework to work and then scaled it up to the full run, but invariably we ran into new problems the greater the scale.
+It therefore took a substantial time to even get this first replication to run.
+It took several months of coding effort to get this correct.
+
 Obviously with complicated diagrams result from the 20,000 runs. The challenge becomes how to visualise these.
@@
 
+So what use is this provenence meta data? Beyond replicability
+
 \begin{figure}
@@
 
+Those with a sharp eye may have noticed the versions of Perl, Python and R were not around when the original experiments were run and obviously this has implications.
+
 \section{Discussion}
 
-What we have learnt from this 
-This is a lot of pfaff. and is still not oven ready
+What we have learnt from this?
+As noted the process of running the experiment with the new provenance and metadata commands was a complete pfaff.
+It involved the repeated running of the experiment with a larger and larger array of inputs until we matched the size of the original experiment.
+There were codes in the both the Python and bash code (calling the Python scripts)
 
-So what use is this provenence meta data. Tracing bad data.
+This has taken me almost a year to get it running.
+Repeating an experiment in a completely automated fashion turns out tbe a really hard task.
+The plan is to adapt this provenance framework to other model running language frameworks, in particular R and Python.
 
@@
 
+The underlying database is probably wrong. We should be using a graphing language such as Gremlin
+
 %
@@
 \bibliographystyle{splncs04}
-\bibliography[keyword={github.com/DougSalt/MABS2023}]{global.bib}
+\bibliography{global.bib}
+% \bibliography[keyword={github.com/DougSalt/SSC2023}]{global.bib}
 %

### 2023-04-19 paper/paper.tex
First version of paper to Gary

@@
 \usepackage{graphicx}
-\usepackage{tabularx}
+\usepackage{longtable}
+\usepackage{xcolor}
+\usepackage{wrapfig}
+
+\newcommand*\ttvar[1]{\texttt{\expandafter\dottvar\detokenize{#1}\relax}}
+\newcommand*\dottvar[1]{\ifx\relax#1\else
+  \expandafter\ifx\string_#1\string_\allowbreak\else#1\fi
+  \expandafter\dottvar\fi}
 % Used for displaying a sample figure. If possible, figure files should
@@
 %
-%\titlerunning{Abbreviated paper title}
+\titlerunning{An automated provenance collection framework}
 % If the paper title is too long for the running head, you can set
@@
 %
-\authorrunning{F. Author et al.}
+\authorrunning{Salt et al.}
 % First names are abbreviated in the running head.
@@
 \begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-15--250 words.
+    We demonstrate a working framework for the automatic recording of
+    provenance and metadata for primarily agent-based models that could easily
+    be adapted to the other modelling environments. We discuss the need for
+    such a framework, the philosophy behind the design we adopted, the
+    implementation, discuss the results and demonstrate a simple tool for for
+    tracing bad data through a provenance graph.
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+\keywords{Provenance \and metadata \and modelling \and automation \and replication.}
 \end{abstract}
@@
 some guidance highlighting the need to keep a notebook of the analysis done and
-a standardised approach to making that notebook; however, the lessons learned
-from the replication exercise in \cite{polhill2017lessons} show that more
+a standardised approach to making that notebook. There are also appear to be
+many scientific workflow tools, examples being, Snakemake
+\cite{koster2012snakemake}, NextFlow \cite{di2017nextflow}, Kepler
+\cite{ludascher2006scientific}, Taverna \cite{hull2006taverna} just to name a
+few. However these are workflow tools, the concentrate on automation and
+repeatability, and if they do record provenance, then it is an after thought.
+It appears that these tools are more centred on automation rather than
+provenance and metadata.
+
+One of the lessons learned from the replication exercise in
+\cite{polhill2017lessons} show that, for the purposes of replicability more
 detailed guidance on the information that should be recorded,  and on tools
@@
 used to explore incentivisation strategies to improve biodiversity
-\cite{polhill_nonlinearities_2013}, \cite{gimona2011exploring}. Belonging to the
+\cite{polhill_nonlinearities_2013,gimona2011exploring}. Belonging to the
 ‘typification’ class of social simulations  \cite{boero2005does}.  This work
 involved the analysis of the outputs of approximately 20,000 runs of the model
-using a number of techniques aimed at demonstrating nonlinearities in the
+using a number of techniques aimed at demonstrating non-linearities in the
 relationship between incentivisation and biodiversity outcome.  Recording
@@
 currently there are no codified standards as to how this should be done for
-ABMs. For FEARLUS-SPOMM, the methods used drew heavily on statistical
+agent-based models. For FEARLUS-SPOMM, the methods used drew heavily on statistical
 techniques available as R packages that are as part of core R functionality.
@@
 work involved great deal of exploration of different ways of attempting to
-visualise and analyse the nonlinearities in the model results, not all of
+visualise and analyse the non-linearities in the model results, not all of
 which were likely to be reported in the paper. Such logs are therefore not the
@@
 analysed used a mixture of text formats, some Perl scripts were also written to
-process that output into a CSV file for easy import into R. When the MIRACLE pr
-ject \cite{parker2019final} provided a context in which the replication of that
-analysis was necessary, an opportunity was created to test the viability of the
-above strategy.
+process that output into a CSV file for easy import into R. When the MIRACLE
+project \cite{parker2019final} provided a context in which the replication of
+that analysis was necessary, an opportunity was created to test the viability
+of the above strategy.
 
-In the rest of this paper, we describe  a tool for automatically recording
-metadata, which can be incorporated into the analysis replication process, and
-how this  was used to regenerate some of the figures in
-\cite{polhill_nonlinearities_2013} and we test a tool to support keeping the
-records needed to per- form replication more easily, before making some
-recommendations in concluding remarks.
+If we make the assumptions that the input data is accurate and relevant; the
+model bug-free and rigorously correct then the manifest problem with
+agent-based modelling is the number of replications required to provide
+adequate probability that an interesting result is not the null hypothesis, and
+just down to a matter of chance. Realistically this involves many, many
+repetitions of the same experiment or small variations on the parameters for a
+given experiment. Similarly infeasible is the manual running of these many
+experiments and thus most resort to some kind of automation. We refer to this
+as the automatic and replicability problem. In  addition to this there is the
+problem of recording which data is being used, which data is being produced,
+how it was gathered or created and how the resultant data was processed. Given
+20,000 experiments the amount of such data is not inconsiderable. Moreover this
+data needs to be in a form that is analysable and visualisable.
+
+The first of these two problems, the automation and replicability of
+experiments  may be and as a matter of course and routine is already solved by
+\emph{scripting}.  In the context of computing scripting is usually a program
+or sequence of instructions carried out by another program rather than the
+processor itself. This means scripting languages may be run by an
+operating system, for instance, rather than compiled down to machine code
+running directly on a processor.  Scripting languages work in one of two ways.
+Either a program is is translated to  byte code, each code represents an
+instruction to a \emph{virtual machine}.  This may be thought of as a software
+computer running on the computer, and has the  obvious advantage that as long
+as the byte code does the same thing everywhere, then the code written in this
+language should be completely agnostic as to the hardware or operating system
+it is running on.  Examples of this type of approach are Python, Java, Julia,
+Perl and  C\#. In particular NetLogo uses the Java virtual machine. These
+languages are very popular, as not only are they portable, they do not require
+the significant overhead in terms of configuration that truly compiled
+languages such as C or C++ require.  Most of the compilation and linking
+process, being computationally intensive and historically slow  has already
+taken place in the virtual machine. A modeller can write code once and it
+should run anywhere capable of running that the virtual machine for which the
+code was originally written. Because of this these scripting languages tend to
+be hardware and operating system agnostic (meaning the code does not 'care'
+where it is run) and more importantly fast to write and share.
+
+More immediately relevant to the problem of automation and replicability, the
+other form of scripting language is a program consisting of a series of
+keywords which trigger sets of conveniently grouped, parameterised
+instructions. A program, known as the \emph{the shell}, in Unix or Linux
+terminology interprets these keywords and their parameters to the more arcane
+instruction set required by the underlying processor to perform some
+computation, such as invoking an executable machine code model. These shells
+are pre-compiled parsers, so were historically limited in terms of their
+expressibility and vocabulary size. Hence, they are generally less expressive
+than the virtual machines discussed above, but are still Turing complete. This
+means the coding of complex algorithms can be achieved but is awkward and slow
+because the keywords are not optimised for general purpose computation, but for
+carrying out operations at level of the operating system. For example, file
+control, with atomicity at the granularity of a single file can be done very
+efficiently using these kinds of languages. Because the latter is effectively
+utilising the operating system of the computer directly, then this grouping of
+languages is generally used to control an existing computer. That is, they are
+used as \emph{job control languages} (JCL). Examples of such languages are
+Bash, Ash, Dash, zsh, csh, Windows Power Shell, MSDOS, etc. Beyond the fact
+that scripting provides automation for large experiment sets, it also allows a
+certain degree of replicability, in that you to reproduce the experiment, then
+all that is required is that the scripts are rerun. However this is
+repeatability rather than replication. The aim is to just to rerun the
+experiment, rather than reproduce a given set of results.  We believe that
+anybody who does serious agent-based modelling experimentation should at the
+very least be scripting most of their experiments. In our experience the
+preparation and the execution of the experiment is relatively straight-forward
+to script. Post processing to the final results of data, diagrams, etc., tends
+to be a little neglected in the hurry to publish results.  Scripting is usually
+(although not exclusively) a text based activity, which means it is amenable to
+being programmatically produced.
+
+The second problem of recording of data, we decided on the use of a relational
+database. This was selected on the basis that such technologies have a proven
+track record of recording and manipulating reasonably large volumes of data.
+This is an established technology. It is also portable and many proven
+libraries exist for the manipulation and use of such databases. And it is this
+context that the discussion of scripting languages and virtual machines becomes
+relevant. The approach we utilised when manipulating this database was a
+scripting language that utilised a virtual machine, specifically Python. As
+mentioned this gives a portability allowing us to develop on laptops and run
+the resultant code in high-performance computing environments with little, or
+no reconfiguration.
+
+Thus the solution we selected was to use a proper JCL, in this case Bash to
+provide the necessary automation and replication models and call the provenance
+functions, written in a popular scripting language (Python) to update a
+standard relational database. In this case we selected Sqlite3 for local
+development and PostgreSQL for more heavy lifting in a distributed environment
+such as in high-performance infrastructure. Thus we have created a provenance
+tool which is based purely on scripting tools and a standard kind of database.
+
+\begin{wrapfigure}{l}{0.5\textwidth} 
+    \includegraphics[width=0.48\textwidth]{img/schema.jpeg}
+\caption{SSREPI Schema} \label{fig:schema} \end{wrapfigure}
+
+Since the inception of this project, we have subsequently uncovered a similar
+project which specifically concentrates on provenance data. This is the FAIR
+data pipeline \cite{mitchell2022fair} in which the provenance is automatically
+recorded by embedded code in R, C++, Java, Julia and Python.  One of the aims
+was not to modify the original code, i.e. the original experiment as unchanged
+as possible. That is we wanted to change the framework code rather than the
+core code. The core code in the example above being the executable, pre-compiled
+in Objective-C \cite{objectivec}. Additionally the support code, i.e. the code
+that prepared the data, a Perl \cite{perl563} script and the code that did the
+post analysis, a mixture of Perl and R \cite{R422}.  This meant we developed
+the database access in Python \cite{python368}, wrote wrapper scripts in Bash
+\cite{bash4420}. This differs in approach to [ibid] which is embedded in the
+code that constitutes the body of the experiment. Moreover our framework
+generalises to more more source metadata (paper, data, other experiments), and
+sis written with the requirements of agent-based modelling first and foremost.
+That is not to say that we do not have similar ambitions to automatically place
+provenance tracing code automatically embedded in experimental model code.
+
+The rest of this paper, we describe the scripting tool we have developed  for
+automatically recording metadata, which can be incorporated into the analysis
+replication process, and how this  was used to regenerate some of the figures
+in \cite{polhill_nonlinearities_2013}.
+In addition we show a simple tool we have already developed to trace data through the provenance graph, and as usual suggest additional work we would like to pursue and other ideas.
 
@@
 Social Simulation REplication Interface or SSREPI. This is the schema shown in
-\ref{fig:schema}.  This schema has evolved since it original conception.  The
-newest version of this document may be found here \cite{polhill2022miracle}. This is a database
-schema derived from the Dublin Core \cite{weibel2000dublin}, the standard XML
-datatypes \cite{biron2004xml} and the PROV-O ontologies \cite{missier2013w3c}.
-The schema is designed with agnosticisim towards the underlying database
-technology and as such has been implmented both in PostgresSQL
-\cite{stonebraker1991postgres} and Sqlite3 \cite{sqliteorg2023syntax}.
+figure \ref{fig:schema}.  This schema has evolved since it original conception.
+The newest version of this document may be found here
+\cite{polhill2022miracle}. This is a database schema derived from the Dublin
+Core \cite{weibel2000dublin}, the standard XML datatypes \cite{biron2004xml}
+and the PROV-O ontologies \cite{missier2013w3c}.  The schema is designed with
+agnosticisim towards the underlying database technology and as such has been
+implemented both in PostgresSQL \cite{stonebraker1991postgres} and Sqlite3
+\cite{sqliteorg2023syntax}.
 
-One of the aims was to leave as much of the code, i.e. the original experiment
-as unchanged as possible. That is we wanted to change the framework code rather
-than the core code. The core code in the example above being the executable,
-precompiled in Objective-C \cite{objectivec}. Additionally the support code, i.e. the
-code that prepared the data, a Perl \cite{perl563} script and the code that did the
-post analysis, a mixture of Perl and R \cite{R422}.  This meant we developed the
-database access in Python \cite{python368}, wrote wrapper scripts in Bash \cite{bash4420}. This
-differs in approach to FAIR data piperline \cite{mitchell2022fair} in which the provenance framework is
-embedded in the code that constitutes the body of the experiment as it were.
+There are two important dimensions of distinction to this schema. First is
+fine- versus coarse- grained metadata. Second is provenance versus workflow.
+Coarse-grained metadata describes how particular files come (or came) into
+being, or were (or could be) used to bring other files into being. Fine-grained
+metadata describes specific values recorded in social simulation outputs. To
+make the distinction concrete, suppose a simulation produces a CSV file. The
+data within the CSV file are covered by fine-grained metadata, whilst the fact
+that the simulation produces the CSV file is coarse-grained. Turning to the
+other dimension, provenance metadata describes what actually happens (run W of
+simulation X produced output file Y), whilst workflow metadata describes what
+could happen (simulation X produces an output file of type Z). The distinctions
+are summarised in \ref{fig:finegrain} .
 
-With the tool in the current state this involved labouriously going through the
-code and inserting the some of the following Bash-based commands, shown in
+\begin{wrapfigure}{r}{0.5\textwidth}
+\includegraphics[width=0.5\textwidth]{img/fine-grain-vs-coarse-grain.jpeg}
+\caption{Fine grain vs coarse grain provenance} \label{fig:finegrain}
+\end{wrapfigure}
+
+Each table in schema shown in fig. \ref{fig:schema} was coded as an object type
+in Python. Each table row was represented as an instance of such an object.
+This design approach was adopted to enforce a consistent coding methodology
+across all tables.
+
+This then just left the interface between the two scripting languages. The
+design decision was taken to do domain specific processing in the Bash
+scripting language. This was done for due to the repetitive and time consuming
+nature of system testing: amendments in the job control language being easier
+to do quickly. This is a decision that may need revisiting. The problem being
+that Bash invocations are computationally and time expensive. Complex
+processing should be abstracted away from the slower language, in this case job
+control language to improve speed and remove what might be regarded as
+inappropriate computation for such an environment. Consequently as the system
+stands the commands that coordinated between Bash and Python are few and
+simple, the harder processing having been pushed down into the job control
+level, i.e. in the Bash scripts.
+
+These coordinating commands are:
+
+\begin{itemize} \item {\color{green} \ttvar{create_database.py}} - creates a
+        database idempotently.  \item {\color{green} \ttvar{exists.py}} - checks
+        if a particular row in a table exists.  \item {\color{green}
+            \ttvar{get_value.py}} - gets any specified single value from a
+            table given the primary key.  \item {\color{green}
+\ttvar{get_values.py}} \item {\color{green} \ttvar{next_study.py}} \item
+{\color{green} \ttvar{update.py}} - idempotently updates a particular row in a
+table.  \end{itemize}
+
+Note the use of idempotent, indicating that multiple operations on
+the same entity will leave it unchanged after the initial operation.  For
+example multiplying something by 1 is idempotent. This allows for a less
+rigorous exception criteria, for instance if a row or entity already exists.
+This implies that initialisation must be performed with care, as older data
+will not necessarily be destroyed but overwritten with newer values, but
+also retaining any values that already exist. Again this design decision was
+expedient on rapid development and may need revisiting. However the schema is
+and its interaction is quite complex and this would involve critical path
+analysis
+
+So each of the Bash commands listed in table \ref{tab:commmands} composes over
+the above python commands to populate the SSREPI database in a consistent and
+logical manner.
+
+For the purposes of this short paper we are concentrating on a demonstration of
+recording provenance. Indeed we use the example, mentioned in the introduction
+of \cite{polhill2022miracle}, and modified the original Bash \cite{bash4420}
+scripts to include the what we denote as \textit{primitives}.  In order to
+record provenance we coded the following primitives from the SSREPI interface
+definition \cite{polhill2022miracle}. As noted above these are implemented in a
+mixture of Python and Bash.
+
+\tiny \begin{longtable}{|l|p{2cm}|p{7cm}|} \hline Primitive & Type & Purpose \\
+    \hline {\color{blue} \ttvar{SSREPI_require\_minimum}} & Metadata & Lower
+    bound on software hardware required \\ {\color{blue}
+    \ttvar{SSREPI_require\_exact}} & Metadata & Exact bound on software
+    hardware required \\ {\color{blue} \ttvar{SSREPI_application}} & Provenance
+    \& Metadata & specifies some executable \\ {\color{blue} \ttvar{SSREPI_me}}
+    & Provenance \& Metadata & Determines executable being run or returns a
+    proper reference to the executable being run. \\ {\color{blue}
+    \ttvar{SSREPI_run}} & Provenance \& metadata & Blocking invocation of an
+    executable which will allow the specification of all inputs, outputs and
+    arguments. Creates run-time provenance information as well \\ {\color{blue}
+    \ttvar{SSREPI_batch}} & Provenance \& metadata & Non blocking invocation of
+    an executable which will allow the specification of all inputs, outputs and
+    arguments. Creates run-time provenance information as well \\ {\color{blue}
+    \ttvar{SSREPI_argument}} & Provenance & An argument type to an exectuable
+    \\ {\color{blue} \ttvar{SSREPI_output}} & Provenance & An output type from
+    an executable \\ {\color{blue} \ttvar{SSREPI_input}} & Provenance & An
+    input type for an exectuable \\ {\color{blue}
+    \ttvar{SSREPI_hutton\_person}} & Metadata & Uses our institutions databases
+    to populate metadata for a given individual \\ {\color{blue}
+    \ttvar{SSREPI_person}} & Metdata & Provide metadata for a particular actor
+    within this system\\ {\color{blue} \ttvar{SSREPI_project}} & Metadata &
+    Specifies a project which contains all studies \\ {\color{blue}
+    \ttvar{SSREPI_study}} & Metadata & A set of experiments makes up a single
+    study \\ {\color{blue} \ttvar{SSREPI_set}} & Metdata & Sets the default
+    licence and other metadata \\ {\color{blue} \ttvar{SSREPI_involvement}} &
+    Metadata & Links personnel to a study \\ {\color{blue}
+    \ttvar{SSREPI_paper}} & Metadata & A paper associated with this study \\
+    {\color{blue} \ttvar{SSREPI_make_tag}} & Metadata & Used for building a
+    folksonomy \\ {\color{blue} \ttvar{SSREPI_tag}} & Metadata & Used to tag
+    any entity with a folksonomy tag \\ {\color{blue}
+    \ttvar{SSREPI_contributor}} & Metadata & A  person with some kind of
+    relation to an executable or script. \\
+
+    {\color{blue} \ttvar{SSREPI_statistical_method}} & Metadata & A statistical
+    method is an approach to computing some statistics. It may be implemented
+    in or as part of an application. A statistical method generates one or more
+    statistical variables as its results, and may use the results of another
+    statistical method in its computation. For example, computing the standard
+    deviation of some data uses the mean of those data. \\ {\color{blue}
+    \ttvar{SSREPI_visualisation}} & Metadata & A visualisation is the process
+    of creating an image to depict one or more (typically more than one)
+    values. The results of a visualisation appear in a container.\\
+    {\color{blue} \ttvar{SSREPI_statistics}} & Metadata & Statistics are
+    activities that compute  and populate the values of statistical rvariables.
+    They operate on raw data that are retrieved from the values using a query.
+    To replicate a set of statistics, the query can be rerun, selecting values
+    that are pointed to by containers entries \\ {\color{blue}
+    \ttvar{SSREPI_visualisation_method}} & Metadata & This describes methods
+    for generating visualisations, which then may appear in the content of a
+    container produced by a process running an application that implements it.
+    \\
+
+    {\color{blue} \ttvar{SSREPI_implements}} & Metadata & Links a statistical
+    or visualisation method to an application \\ {\color{blue}
+    \ttvar{SSREPI_parameter}} & Metadata & A Parameter is the name of a
+    parameter taken by a statistical or visualisation method, used to configure
+    the way it behaves. \\ {\color{blue} \ttvar{SSREPI_statistical_variable}} &
+    Metadata &  A name for (one of) the result(s) of a statistical method. \\
+    {\color{blue} \ttvar{SSREPI_visualisation_variable}} & Provenance \&
+    Metadata & Declares a named variable of interest \\ {\color{blue}
+    \ttvar{SSREPI_variable}} & Metadata & Names a variable of interest \\
+    {\color{blue} \ttvar{SSREPI_statistical_variable_value}} & Provenance \&
+    Metdata & Sets an actual value for a named statistical variable \\
+    {\color{blue} \ttvar{SSREPI_value}} & Provenance & Sets an actual value.
+    This can be connected to any metadata value such         \\ {\color{blue}
+    \ttvar{SSREPI_content}} & Metadata & Links a kind of output/input/argument
+    to a variable  \\ {\color{blue} \ttvar{SSREPI_person_makes_assumption}} &
+Metadata & links a person to an assumption \\ \hline \caption{Available Bash
+commands for provenance and metadata gathering} \label{tab:commands}
+\end{longtable}
+
+\normalsize
+
+Broadly speaking {\color{blue} \ttvar{SSREPI_application}}, {\color{blue}
+\ttvar{SSREPI_run}}, {\color{blue} \ttvar{SSREPI_batch}}, {\color{blue}
+\ttvar{SSREPI_input}}, {\color{blue} \ttvar{SSREPI_output}} and {\color{blue}
+\ttvar{SSREPI_argument}} are responsible for recording coarse grain provenance.
+{\color{blue} \ttvar{SSREPI_value}}, {\color{blue}
+\ttvar{SSREPI_visualisation_variable_value}}, {\color{blue}
+\ttvar{SSREPI_statistical_variable_value}}, {\color{blue} \ttvar{SSREPI_run}}
+and {\color{blue} \ttvar{SSREPI_batch}} record fine-grain provenance. The
+remaining primitives are largely about recording metadata.
+
+With the tool in the current state this involved laboriously going through the
+code and inserting the some of the Bash-based commands, shown in
 table \ref{tab:commands}. These commands have a fairly complex parameter set as
 it stands, and no way of checking whether the code worked other than running
-it, and this led to one of the immediate problems with this method.
+it, initially on a small subset of the experiment and then scaling up. This was
+extremely time-consuming.
 
-\begin{figure}
-\includegraphics[width=\textwidth]{img/schema.jpeg}
-\caption{SSREPI Schema} \label{fig:schema}
-\end{figure}
+Elaboration of all this may be found in the man pages and design documents in
+the public repository at \url{https://github.com:/DougSalt/ABM-metadata}.
 
-There are two important dimensions of distinction to this schema. First is
-fine- versus coarse- grained metadata. Second is provenance versus
-workflow. Coarse-grained metadata describes how particular files come (or
-came) into being, or were (or could be) used to bring other files into
-being. Fine-grained metadata describes specific values recorded in social
-simulation outputs. To make the distinction concrete, suppose a simulation
-produces a CSV file. The data within the CSV file are covered by
-fine-grained metadata, whilst the fact that the simulation produces the
-CSV file is coarse-grained. Turning to the other dimension, provenance
-metadata describes what actually happens (run W of simulation X produced
-output file Y), whilst workflow metadata describes what could happen
-(simulation X produces an output file of type Z). The distinctions are
-summarised in \ref{fig:finegrain} .
+\begin{figure} \includegraphics[width=\textwidth]{img/workflow.pdf}
+\caption{The workflow sub-graph} \label{fig:workflow} \end{figure}
 
-\begin{figure}
-\includegraphics[width=\textwidth]{img/fine-grain-vs-coarse-grain.jpeg}
-\caption{Fine grain vs coarse grain provenance} \label{fig:finegrain}
-\end{figure}
+We developed a very basic scheduler for use in an high performance computing
+environment, where there are many resources available both on the local
+machine. In particular for the situation where they is a single machine that
+has a large memory and many cores, the non-blocking version of the  two
+execution primitives, specifically {\color{blue} \ttvar{SSREPI_batch}}
+implements a kind of a simple scheduler locally using the sub-processing
+commands available to Unix-like systems. This was an historical solution to
+schedule in an high performance computing environment that existed prior to
+clustered computing situations. Clustering is the utilisation of one or more
+instances of such powerful computing machinery. Generally schedulers have been
+developed for such environments, as such SGE \cite{gentzsch2001sun} and the one
+we use as default, Slurm \cite{yoo2003slurm}. Scheduling is now largely handled
+outwith the framework, which makes the code less complex.
 
-For the purposes of this short paper we are concentrating on a
-demonstration of recording provenance. Indeed we use the example,
-mentioned in the introduction of \cite{polhill2022miracle}, and modified the original Bash \cite{bash4420} scripts to include the what we denote as \textit{primitives} . So an example of the code
-
-In order to record provenance we 
-We coded the following primitives from
-the SSREPI interface definition \cite{polhill2022miracle}. These are implemented in a mixture of Python and Bash.
-
-\begin{table}
-\caption{Table captions should be placed above the
-tables.}\label{tab:commands}
-    \begin{tabularx}{\textwidth}{|l|p{2cm}|p{6cm}|}
-\hline
-Primitive & Type & Purpose \\
-\hline
-        SSREPI\_require\_minimum & Metadata & Lower bound on software hardware required \\ SSREPI\_require\_exact & Metadata & Exact bound on software hardware required \\
-        SSREPI\_application & Provenance \& Metadata & specifies some executable \\
-        SSREPI\_me & Provenance \& Metadata & Determines executable being run or returns a proper reference to the executable being run. \\
-        SSREPI\_run & Provenance \& metadata & Blocking invocation of an executable which will allow the specification of all inputs, outputs and arguments.
-Creates run-time provenance information as well \\
-        SSREPI\_batch & Provenance \& metadata & Non blocking invocation of an executable which will allow the specification of all inputs, outputs and arguments.
-Creates run-time provenance information as well \\
-        SSREPI\_argument & Provenance & An argument type to an exectuable \\
-        SSREPI\_output & Provenance & An output type from an executable \\
-        SSREPI\_input & Provenance & An input type for an exectuable \\
-        SSREPI\_hutton\_person & Metadata & Uses our institutions databases to populate metadata for a given individual \\
-        SSREPI\_person & Metdata & Provide metadata for a particular actor within this system\\
-        SSREPI\_project & Metadata & Specifies a project which contains all studies \\
-        SSREPI\_study & Metadata & A set of experiments makes up a single study \\
-        SSREPI\_set & Metdata & Sets the default licence and other metadata \\
-        SSREPI\_involvement & Metadata & Links personnel to a study \\
-        SSREPI\_paper & Metadata & A paper associated with this study \\
-        SSREPI\_make\_tag & Metadata & Used for building a folksonomy \\
-        SSREPI\_tag & Metadata & Used to tag any entity with a folksonomy tag \\
-        SSREPI\_contributor & Metadata & A  person with some kind of relation to an executable or script. \\
-
-        SSREPI\_statistical\_method & Metadata & A statistical method is an approach to computing some statistics.
-It may be implemented in or as part of an application.
-A statistical method generates one or more statistical variables as its results, and may use the results of another statistical method in its computation.
-For example, computing the standard deviation of some data uses the mean of those data.
-\\
-        SSREPI\_visualisation & Metadata & A visualisation is the process of creating an image to depict one or more (typically more than one) values.
-The results of a visualisation appear in a container.\\
-        SSREPI\_statistics & Metadata & Statistics are activities that compute  and populate the values of statistical rvariables.
-They operate on raw data that are retrieved from the values using a query.
-To replicate a set of statistics, the query can be rerun, selecting values that are pointed to by containers entries \\
-        SSREPI\_visualisation\_method & Metadata & This describes methods for generating visualisations, which then may appear in the content of a container produced by a process running an application that implements it.
-\\
-
-        SSREPI\_implements & Metadata & Links a statistical or visualisation method to an application \\
-        SSREPI\_parameter & Metadata & A Parameter is the name of a parameter taken by a statistical or visualisation method, used to configure the way it behaves. \\
-        SSREPI\_statistical\_variable & Metadata &  A name for (one of) the result(s) of a statistical method. \\
-        SSREPI\_visualisation\_variable & Provenance \& Metadata & Declares a named variable of interest \\
-        SSREPI\_variable & Metadata & Names a variable of interest \\
-        SSREPI\_statistical\_variable\_value & Provenance \& Metdata & Sets an actual value for a named statistical variable \\
-        SSREPI\_value & Provenance & Sets an actual value. This can be connected to any metadata value such         \\ 
-        SSREPI\_content & Metadata & Links a kind of output/input/argument to a variable  \\
-        SSREPI\_person\_makes\_assumption & Metadata & links a person to an assumption \\
-\hline
-\end{tabularx}
-\end{table}
-
-Broadly speaking SSREPI\_application, SSREPI\_run, SSREPI\_batch, SSREPI\_input,
-SSREPI\_output and SSREPI\_argument are responsible for recording coarse grain
-provenance. SSREPI\_value, SSREPI\_visualisation\_variable\_value, SSREPI\_statistical\_variable\_value, SSREPI\_run and SSREPI\_batch record fine-grain provenance. The remaining
-primitives are largely about recording metadata.
-
-Elaboration of all this may be found in the man pages and design documents in the public repository at git@github.com:DougSalt/ABM-metadata.git
-
-\begin{figure}
-\includegraphics[width=\textwidth]{img/workflow.pdf}
-\caption{The workflow sub-graph} \label{fig:workflow.pdf}
-\end{figure}
+{\color{blue} \ttvar{SSREPI_run}}, the other executing primitive will block
+until the command, code or model is has invoked has completed. This is
+necessarily used in single threaded coordination context, such as gathering all
+outputs necessary before proceeding to the next stage of processing. For
+instance, initial post processing of an all experiments would represent such a
+single threaded context.
 
@@
 
-As mentioned the method undertaken in this paper involved inserting the correct
-Bash-based commands, listed in table \ref{tab:commands}. These commands
-have a fairly complex parameter set as it stands, and no way of checking
-whether the code worked other than running it, and this led to one of the
-immediate problems with this method. The only way to see if the code was
-working was to run it, a suck it and see methodology. This was extremely time consuming. We took the obvious approach and ran small subsets of
-the ~20,000 runs or so, to see if we could get the framework to work and then scaled it up to the full run, but invariably we ran into new problems the greater the scale.
-It therefore took a substantial time to even get this first replication to run.
-It took several months of coding effort to get this correct.
+The primary result is that the target results and diagrams from the original
+paper \cite{polhill2017lessons} were able to be recreated.  As mentioned the
+method undertaken in this paper involved inserting the correct Bash-based
+commands, listed in table \ref{tab:commands}. These commands have a fairly
+complex parameter set as it stands, and no way of checking whether the code
+worked other than running it. This led to one of the immediate problems with
+this method. The only way to see if the code was working was to run it, a suck
+it and see methodology. This was extremely time consuming. We took the obvious
+approach and ran small subsets of the ~20,000 runs or so, to see if we could
+get the framework to work and then scaled it up to the full run, but invariably
+we ran into new problems the more we increased the scale. It therefore took
+substantial time to even get this first replication to run. It took several
+months of coding effort to get this correct. Generally it would alternate
+between parameter errors for the calls in table \ref{tab:commands}, and errors in the scripts making calls to record the provenance data to the database.
+Each iteration would take weeks to run, before failure as we increased the scale.
 
-Obviously with complicated diagrams result from the 20,000 runs. The challenge becomes how to visualise these.
-There are several graphs produced so far, these being
+After a run had completed successfully the database was run through several
+programs to produce the sub-graphs detailed in fig. \ref{fig:schema}. These
+programs are capable of producing the following subgraphs:
 
-We show the workflow graph in fig. \ref{fig:workflow} and a section of the
-proveance graph in fig. \ref{fig:sub-provenance}
+\begin{itemize} \item \textbf{Analysis} - fine grain provenance pertaining to
+            statistical and visualisation outputs.
 
-So what use is this provenence meta data? Beyond replicability
+        \item \textbf{Finegrain} - a provenance diagram down to the level of
+        variables.  \item \textbf{Folksonomomy} - a diagram showing annotations
+            against the database, produced and categorised at the discretion of
+            the user doing the annotation \item \textbf{Project} - management
+            metadata. Largest granularity of metadata supported \item
+            \textbf{Provenance} - provenance diagram at the level of file and
+        parameter \item \textbf{Services} - service provided and requirement
+            description \item \textbf{Workflow} - the actual workflow
+\end{itemize}
+
+We developed programs that took the relational data in order to produce "dot"
+files. Such files that can be processed by Graphviz
+\cite{ellson2002graphviz}, and were used to produce the diagrams in figures
+\ref{fig:sub-provenance}, \ref{fig:workflow}, \ref{fig:broken-application} and
+\ref{fig:high-lit-provenance-trace} .
+
+To give an idea of the visualisation available we show the workflow graph in
+fig. \ref{fig:workflow}. This is a section of the workflow and is relative easy
+to follow. The resultant provenance graph is both massive and massively
+complex, given there were 20,000 runs (so 20,000 sets of outputs to record), so we only show a \textit{very small} section of the provenance graph in fig.
+\ref{fig:sub-provenance}.
 
@@
 
-Those with a sharp eye may have noticed the versions of Perl, Python and R were not around when the original experiments were run and obviously this has implications.
+So what use is this provenance metadata? Beyond replicability? Say for instance
+we have a bad dataset. The challenge is not only how to visualise the
+provenance, but to use this data in a useful manner. We propose a simple use
+here.  The "dot" language used by Graphviz constitutes a primitive graph
+database. Indeed there are programs that can transform "dot" files into
+TinkerPop GraphSON format \cite{tinkerpop2015graphson,gv2graphson}. Rather than
+using the relational database data we used the resultant dot files listed above
+to trace bad entities. \footnote{These are directed graphs (unproven as to
+whether they are acyclic - they shouldn't be they are the result of a
+relational database which was designed to be and \textit{should} be acyclic).}
+
+So for an easy to see example we pick the entity {\color{red}\ttvar{Applications.application_3831436655}}, and see how this affects other entities in the graph.
+This is \ttvar{SSS-StopC2-Cluster-create.pl} and we are using the workflow graph to see what might be affected.
+And this is shown in \ref{fig:broken-application}.
+It might be for instance that there is a bug in this program, so we should like to see how the errors might propagate out through the workflow.
+
+\begin{figure} \includegraphics[width=\textwidth]{img/broken-application.pdf}
+\caption{The workflow sub-graph} \label{fig:broken-application} \end{figure}
+
+That example was somewhat contrived and simplistic and gives a flavour.
+However, what if we wanted to trace a bad data set?  If we ignore for this
+original example that the datasets for the above example are generated
+automatically by Perl scripts, and we pick one randomly to illustrate how this
+approach might work. The dataset in question would be:
+{\color{red}\ttvar{Containers.container_42949672955}}, a data file that forms
+a input file to just one experiment.  And here we see, again, a \textit{very
+small} section of of the provenance graph showing what is affected in red. This
+is shown in fig. \ref{fig:high-lit-provenance-trace}
+
+\begin{figure} \includegraphics[width=\textwidth]{img/high-lit-provenance-trace.pdf}
+\caption{The workflow sub-graph} \label{fig:high-lit-provenance-trace} \end{figure}
+
+Moreover we can list out the affected entities. In this case, those entities would be:
+
+\tiny
+
+\begin{itemize}
+	\item \ttvar{Applications.application_3450918915}
+	\item \ttvar{Applications.application_648609270}
+	\item \ttvar{Computers.asterix.local}
+	\item \ttvar{Containers.container_1814970370}
+	\item \ttvar{Containers.container_1982026419}
+	\item \ttvar{Containers.container_2050039078}
+	\item \ttvar{Containers.container_2056384913}
+	\item \ttvar{Containers.container_2060874102}
+	\item \ttvar{Containers.container_2387213333}
+	\item \ttvar{Containers.container_2486610989}
+	\item \ttvar{Containers.container_2582525701}
+	\item \ttvar{Containers.container_2759060318}
+	\item \ttvar{Containers.container_2865400753}
+	\item \ttvar{Containers.container_3025688835}
+	\item \ttvar{Containers.container_3307537171}
+	\item \ttvar{Containers.container_3470971297}
+	\item \ttvar{Containers.container_354343442}
+	\item \ttvar{Containers.container_4235735972}
+	\item \ttvar{Containers.container_4294967295}
+	\item \ttvar{Containers.container_441913555}
+	\item \ttvar{Containers.container_505627104}
+	\item \ttvar{Containers.container_800277554}
+	\item \ttvar{Containers.container_878886043}
+	\item \ttvar{Containers.container_900718909}
+	\item \ttvar{Persons.doug}
+	\item \ttvar{Persons.doug_salt}
+	\item \ttvar{Processes.process_23222129888649347509041510897074}
+	\item \ttvar{Processes.process_32647549959702293885495865014666}
+	\item \ttvar{Specifications.R}
+	\item \ttvar{Specifications.bash}
+	\item \ttvar{Specifications.cpus}
+	\item \ttvar{Specifications.disk_space}
+	\item \ttvar{Specifications.memory}
+	\item \ttvar{Specifications.os}
+	\item \ttvar{Specifications.perl}
+	\item \ttvar{Specifications.python}
+	\item \ttvar{Studies.1}
+	\item \ttvar{Users.doug}
+\end{itemize}
+
+\normalsize
 
@@
 
-What we have learnt from this?
-As noted the process of running the experiment with the new provenance and metadata commands was a complete pfaff.
-It involved the repeated running of the experiment with a larger and larger array of inputs until we matched the size of the original experiment.
-There were codes in the both the Python and bash code (calling the Python scripts)
+What we have learnt from this?  As noted the process of running the experiment
+with the new provenance and metadata commands was time-consuming to the point
+of almost being unusable. It involved the repeated running of the experiment
+with a larger and larger array of inputs until we matched the size of the
+original experiment.  This has taken several months of intense work to event get it running. Repeating an experiment in a completely automated fashion turns out to be a really hard task.
 
-This has taken me almost a year to get it running.
-Repeating an experiment in a completely automated fashion turns out tbe a really hard task.
-The plan is to adapt this provenance framework to other model running language frameworks, in particular R and Python.
+However, we have to emphasise that this was the reproduction of the experiment.
+If this framework were available from the start then most of the work would be
+done once the experiments had been completed; that is, the work needed to
+actually repeat the experiment in a consistent and reliable manner. This is the
+key, and it reduces to the same advice that meta-data advocates everywhere
+proclaim, which is to incorporate the metadata from the start and not as an
+after thought.
 
-We demonstrate this by tracing a series of bad data across the diagrams.
+Provenance should be a directed acyclic graph. There may well be loops in the
+schema as it stands. Given sufficient time we can certainly normalise the
+relational database to remove redundancy and the conflict arising from such,
+and once this complete,  we can formally prove the schema to be acyclic.  The
+authors have done their best to normalise it to uncover such cycles, but have
+yet to formally do so. In a schema this complex they may well still exist.
+Until the schema is formally analysed, then we can not say with confidence that
+the schema is actually consistent, in the sense that any recorded provenance is
+not inherently contradictory. Given that such a schema is likely to be
+continually revised as the underlying provenance model changes or additional
+functionality is required, then it would be advisable to automate such proving
+of consistency (assuming correct normalisation).
 
-The proveance should be a directed acyclic graph.
-There may well be loops in the schema as it stands.
-The authors have done their best to normalise it to uncover such cycles, but in a schema this complex they may well still exist.
+In addition to reproducing the experiment automatically, there was some scope
+in utilising the parallelising infrastructure of the high performance computing
+environment.  One of the nice things was that we were able to parallelise part
+of the post processing job, and thus speed up the whole experiment as this was
+proving to be a bit of a bottleneck. However, careful thought needs to be given
+to this. In the case of agent-based modelling such stochasticism in inherent to
+the testing methodology, and it would be expected that such changes should have
+little impact. For other testing frameworks, particular those that do not rely
+on massive replication, then such changes may well have unintended
+consequences.  Indeed, those with a sharp eye may have noticed the versions of
+Perl, Python and R we use were not around when the original experiments were
+run and obviously this has implications. However, the framework does offer the
+ability to record and enforce metadata about programming versions.
 
-Eventually we should like to take several databases run some machine learning across them to see if there are any commonalities in experiment set up and post processing.
+The underlying database is almost certainly unsuitable. We should probably be
+using a graphing language such as Gremlin \cite{rodriguez2015gremlin}. The
+advantages of using graphing databases over relational databases is that such
+languages are inherently designed to store, traverse and query the graphs that
+are the main product of this framework. This makes querying in such languages
+trivial and fast. In a relational databases the SQL statement to do this are
+cumbersome, awkward and probably error prone given their size. Moreover on huge
+datasets they are reportedly slow. Relational databases do have the advantage
+of being an extremely mature technology and the availability of utilities that
+implies. As mentioned, the advantage of using the relational database is that
+structured query language is standard for all such databases, and therefore
+queries written for any relational database should be the same. A tool we
+already use, Graphviz already does act like a graphing database language to a
+certain extent. Indeed we use the Graphviz "dot" files, rather than the
+relational schema in our simple tools tools to trace bad data through our
+provenance graphs.  However the "dot" language of Graphviz is primarily
+purposed as a diagramming language and lacks the sophistication, such as a
+built-in query language of say, Gremlin.
 
-Additionally we want to produce modules in other languages as well, such as R, python and Julia. Especially the latter.
+For future we work, we should eventually like to take many provenance databases run some machine learning across them to see if there are any commonalities.
+We hope to uncover similarities in setup, execution and post-processing that could form core, reusable and proven components for primarily agent-based modelling but also for other modelling frameworks.
 
-One of the nice things was that I was able to parallelise the post processing job simply.
+The plan is to adapt this provenance framework to other model running language
+frameworks, in particular R, Python, Julia, Java and thence NetLogo. This would
+take the approach of the FAIR data pipeline \cite{mitchell2022fair}, but unlike
+the example we have presented here such provenance would be embedded in the
+experiments as a matter of course. We were constrained by the nature of the
+replication here to use Bash as our means of recording provenance. However this
+does show that such a provenance framework may be "retrofitted" to existing
+experiments (although possibly at the cost of the programmer's sanity given the
+minutiae involved). We have tentative potential adopters using R and Python as their primary modelling language.
 
-The underlying database is probably wrong. We should be using a graphing language such as Gremlin
-
-%
 % ---- Bibliography ----

### 2023-04-19 paper/paper.tex
Updates from Overleaf

@@
 %
-\title{Towards automated provenance collection for experimental runs of agent-based models\thanks{Supported by The Scottish Government}}
+\title{Towards automated provenance collection for experimental runs of agent-based models\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}}
+%%% GP: if you ever want the standard acknowledgement text for C5, it's on our
+%%% GP: project webpage at https://large-scale-modelling.hutton.ac.uk/
 %
@@
 %
-\author{Doug Salt\inst{1}\orcidID{0000-1111-2222-3333} \and
-Gary Polhill\inst{2,3}\orcidID{1111-2222-3333-4444} }
+\author{Doug Salt\inst{1}\orcidID{0000-0001-5186-9388} \and
+Gary Polhill\inst{1}\orcidID{0000-0002-8596-0590} }
+
+%%% GP: to discuss -- which other authors should we include here?
+%%% GP: Kit Macleod? (Has he had intellectual input into this via WP1?)
+%%% GP: Coran Musk? (Random acknowledgement for looking after the machines)
+%%% GP: Lorenzo Milazzo? (As the original SSREPI developer)
+%%% GP: Dawn Parker? (As MIRACLE PI -- do we see this as a MIRACLE paper?)
 %
-\authorrunning{Salt et al.}
+\authorrunning{D. Salt et al.}
 % First names are abbreviated in the running head.
@@
 %
-\institute{The James Hutton Institute, Craigiebuckler Aberdeen AB15 8QH Scotland
-\email{doug.salt@hutton.ac.uk}\\
-\url{http://hutton.ac.uk}
+\institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
+\email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
+\url{https://www.hutton.ac.uk/}
 }
@@
 Replication of social simulation results has been highlighted as a significant
-issue for the ABM for a number of years (e.g.  \cite{edmonds2003replication}),
-and in particular the paper that forms the basis of this work
-\cite{polhill2017lessons}.  The focus of replication work has previously just
-been on the model itself, but as was shown in ibid, the analysis of the
-outputs of the model can potentially be just as complex, and no less difficult
-to replicate unless adequate records are kept. Indeed there as an additional
-attempt to reproduce the results and even with the lessons learnt from ibid, it
-still took months to recreate the final diagrams that were submitted to paper.
-The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
+issue for the ABM community for a number of years (e.g. \cite{edmonds2003replication}).
+The paper that forms the basis of this work (\cite{polhill2017lessons}) shows that
+the analysis of the outputs of the model can potentially be just as complex a
+process as developing and running the model itself. Analysis of outputs is no less difficult
+to replicate unless adequate records are kept. The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
 some guidance highlighting the need to keep a notebook of the analysis done and
-a standardised approach to making that notebook. There are also appear to be
-many scientific workflow tools, examples being, Snakemake
+a standardised approach to making that notebook. There are also 
+many scientific workflow tools, such as Snakemake
 \cite{koster2012snakemake}, NextFlow \cite{di2017nextflow}, Kepler
-\cite{ludascher2006scientific}, Taverna \cite{hull2006taverna} just to name a
-few. However these are workflow tools, the concentrate on automation and
-repeatability, and if they do record provenance, then it is an after thought.
-It appears that these tools are more centred on automation rather than
-provenance and metadata.
+\cite{ludascher2006scientific} and Taverna \cite{hull2006taverna}. However, since
+these are workflow tools, the focus is on automation and
+repeatability rather than provenance, which, if it is included, is as an afterthought.
 
 One of the lessons learned from the replication exercise in
-\cite{polhill2017lessons} show that, for the purposes of replicability more
-detailed guidance on the information that should be recorded,  and on tools
-that could be used to support the process. Ideally the aim should be to
-completely automate the process, record accurately sources of data used,
-version check applications used essentially providing a complete graph from
-data to result.
+\cite{polhill2017lessons} was that, for the purposes of replication, more
+detailed guidance on the information that should be recorded is needed. Since
+recording such metadata is tedious (and error-prone) for humans, any such
+guidance should be accompanied by specification of tools
+that could be used to support the process. In the ideal world, the process
+of recording provenance metadata would be completely automated, essentially providing a complete graph from data (including their sources) through applications (model, scripts and analysis tools, including versions thereof) to result.
 
@@
 \cite{polhill_nonlinearities_2013,gimona2011exploring}. Belonging to the
-‘typification’ class of social simulations  \cite{boero2005does}.  This work
+‘typification’ class of social simulations  \cite{boero2005does}, this work
 involved the analysis of the outputs of approximately 20,000 runs of the model
-using a number of techniques aimed at demonstrating non-linearities in the
+using a number of techniques aimed at demonstrating nonlinearities in the
 relationship between incentivisation and biodiversity outcome.  Recording
-workflow data on the process used to create analysis can be challenging, and
+provenance metadata on the process used to analyse and visualise the outputs is challenging, and
 currently there are no codified standards as to how this should be done for
-agent-based models. For FEARLUS-SPOMM, the methods used drew heavily on statistical
-techniques available as R packages that are as part of core R functionality.
+agent-based models. For FEARLUS-SPOMM, the analysis and visualisation methods used drew heavily on statistical
+techniques available as R packages.
 Although R allows transcripts of interactive terminal sessions to be saved, the
-work involved great deal of exploration of different ways of attempting to
-visualise and analyse the non-linearities in the model results, not all of
-which were likely to be reported in the paper. Such logs are therefore not the
-best way to record the means by which the outputs were analysed, and hence the
+work involved great deal of exploration of alternative analyses and visualisations, not all of
+which were reported in the manuscript as finally accepted. Such transcripts are therefore not the
+best way to record the means by which a model's outputs were analysed, and hence the
 strategy used was to save each analysis or visualisation in a(n R) script.
-Since the output from the (Swarm) software that generated the output data being
-analysed used a mixture of text formats, some Perl scripts were also written to
-process that output into a CSV file for easy import into R. When the MIRACLE
+Since the output from the (Swarm) model software used a mixture of text formats, some Perl scripts were also written to
+process that output into CSV format for easy import into R. When the MIRACLE
 project \cite{parker2019final} provided a context in which the replication of
 that analysis was necessary, an opportunity was created to test the viability
-of the above strategy.
+of the strategy of relying on scripts to record provenance.
 
-If we make the assumptions that the input data is accurate and relevant; the
-model bug-free and rigorously correct then the manifest problem with
-agent-based modelling is the number of replications required to provide
-adequate probability that an interesting result is not the null hypothesis, and
-just down to a matter of chance. Realistically this involves many, many
-repetitions of the same experiment or small variations on the parameters for a
-given experiment. Similarly infeasible is the manual running of these many
-experiments and thus most resort to some kind of automation. We refer to this
+`Multifinality' (the same initial conditions and parameter settings having qualitatively different results) in agent-based models means we need to ensure that
+reported results are not just down to a matter of chance.
+This is one of many reasons (e.g.
+in empirical contexts especially, calibration, validation and sensitivity analysis) why experiments with ABMs involve large numbers of runs.
+Most ABM experiments therefore resort to some kind of automation, including of the kind provided by workflow tools mentioned above, but also using built-in features of ABM software (e.g.
+in the case of NetLogo, BehaviorSpace).
+We refer to this
 as the automatic and replicability problem. In addition to this there is the

### 2023-04-19 paper/paper.tex
Updates from Overleaf

@@
 Replication of social simulation results has been highlighted as a significant
-issue for the ABM community for a number of years (e.g. \cite{edmonds2003replication}).
+issue for the agent-based modelling (ABM) community for a number of years (e.g. \cite{edmonds2003replication}).
 The paper that forms the basis of this work (\cite{polhill2017lessons}) shows that
@@
 The output analysis replication in this paper concerns earlier work with
-FEARLUS-SPOMM, which is a coupled agent-based model of agricultural decision-
+FEARLUS-SPOMM, which is a coupled ABM of agricultural decision-
 making and species stochastic patch occupancy metacommunity model that has been
@@
 currently there are no codified standards as to how this should be done for
-agent-based models. For FEARLUS-SPOMM, the analysis and visualisation methods used drew heavily on statistical
+ABMs. For FEARLUS-SPOMM, the analysis and visualisation methods used drew heavily on statistical
 techniques available as R packages.
@@
 
-`Multifinality' (the same initial conditions and parameter settings having qualitatively different results) in agent-based models means we need to ensure that
+`Multifinality' (the same initial conditions and parameter settings having qualitatively different results) in ABMs means we need to ensure that
 reported results are not just down to a matter of chance.
@@
 in empirical contexts especially, calibration, validation and sensitivity analysis) why experiments with ABMs involve large numbers of runs.
-Most ABM experiments therefore resort to some kind of automation, including of the kind provided by workflow tools mentioned above, but also using built-in features of ABM software (e.g.
+The tedium (and in larger-scale cases, infeasibility) of conducting each run by hand means most ABM experiments resort to some kind of automation, including of the kind provided by workflow tools mentioned above, but also using built-in features of ABM software (e.g.
 in the case of NetLogo, BehaviorSpace).
+This is fine if all we
+want to do is repeat the same process, but what if we are interested in why a particular instance of that process led to an unexpected outcome?
+Re-executing the workflow will not necessarily generate \textit{exactly} the same result unless we have a record of everything needed to do that (including seeds for pseudo-random number generators).
 We refer to this
-as the automatic and replicability problem. In addition to this there is the
-problem of recording which data is being used, which data is being produced,
-how it was gathered or created and how the resultant data was processed. Given
-20,000 experiments the amount of such data is not inconsiderable. Moreover this
-data needs to be in a form that is analysable and visualisable.
+as the `automation and replication' problem.
+
+%%% GP: Got here as at 2023-04-19T10:59.
 
@@
             table given the primary key.  \item {\color{green}
-\ttvar{get_values.py}} \item {\color{green} \ttvar{next_study.py}} \item
+\ttvar{get_values.py}}  - gets one or more rows given the search values. \item {\color{green}\ttvar{next_study.py}} - gets the next available and unique study number. \item
 {\color{green} \ttvar{update.py}} - idempotently updates a particular row in a
@@
 
-So each of the Bash commands listed in table \ref{tab:commmands} composes over
+So each of the Bash commands listed in table \ref{tab:commands} composes over
 the above python commands to populate the SSREPI database in a consistent and

### 2023-04-19 paper/paper.tex
Added the images that are the basis of the diagrams

@@
 \begin{figure} \includegraphics[width=\textwidth]{img/broken-application.pdf}
-\caption{The workflow sub-graph} \label{fig:broken-application} \end{figure}
+\caption{The workflow sub-graph show the propagation of a broken application (in red) to its outputs} \label{fig:broken-application} \end{figure}
 
@@
 \begin{figure} \includegraphics[width=\textwidth]{img/high-lit-provenance-trace.pdf}
-\caption{The workflow sub-graph} \label{fig:high-lit-provenance-trace} \end{figure}
+\caption{A provenance sub-graph showing (in red) the propagation of bad data through the system} \label{fig:high-lit-provenance-trace} \end{figure}
 

### 2023-05-12 paper/paper.tex
Gary changes

@@
 \usepackage{longtable}
-\usepackage{xcolor}
+\usepackage[usenames,dvipsnames]{xcolor}
 \usepackage{wrapfig}
@@
 %
-\title{Towards automated provenance collection for experimental runs of agent-based models\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}}
+\title{Towards automated provenance collection for experimental runs of agent-based models}
+
+%\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}}
 %%% GP: if you ever want the standard acknowledgement text for C5, it's on our
@@
 %
-\author{Doug Salt\inst{1}\orcidID{0000-0001-5186-9388} \and
-Gary Polhill\inst{1}\orcidID{0000-0002-8596-0590} }
+% \author{Doug Salt\inst{1}\orcidID{0000-0001-5186-9388} \and
+% Gary Polhill\inst{1}\orcidID{0000-0002-8596-0590} }
 
@@
 %
-\authorrunning{D. Salt et al.}
+%\authorrunning{D. Salt et al.}
 % First names are abbreviated in the running head.
@@
 %
-\institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
-\email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
-\url{https://www.hutton.ac.uk/}
-}
+% \institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
+% \email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
+%\url{https://www.hutton.ac.uk/}
+%}
 %
@@
 
-The first of these two problems, the automation and replicability of
-experiments  may be and as a matter of course and routine is already solved by
-\emph{scripting}.  In the context of computing scripting is usually a program
-or sequence of instructions carried out by another program rather than the
-processor itself. This means scripting languages may be run by an
-operating system, for instance, rather than compiled down to machine code
-running directly on a processor.  Scripting languages work in one of two ways.
-Either a program is is translated to  byte code, each code represents an
-instruction to a \emph{virtual machine}.  This may be thought of as a software
-computer running on the computer, and has the  obvious advantage that as long
-as the byte code does the same thing everywhere, then the code written in this
-language should be completely agnostic as to the hardware or operating system
-it is running on.  Examples of this type of approach are Python, Java, Julia,
-Perl and  C\#. In particular NetLogo uses the Java virtual machine. These
-languages are very popular, as not only are they portable, they do not require
-the significant overhead in terms of configuration that truly compiled
-languages such as C or C++ require.  Most of the compilation and linking
-process, being computationally intensive and historically slow  has already
-taken place in the virtual machine. A modeller can write code once and it
-should run anywhere capable of running that the virtual machine for which the
-code was originally written. Because of this these scripting languages tend to
-be hardware and operating system agnostic (meaning the code does not 'care'
-where it is run) and more importantly fast to write and share.
+%%% GP: Restarting 2023-05-11T15:36 -- these details on scripts can go and I am about to delete them
 
-More immediately relevant to the problem of automation and replicability, the
-other form of scripting language is a program consisting of a series of
-keywords which trigger sets of conveniently grouped, parameterised
-instructions. A program, known as the \emph{the shell}, in Unix or Linux
-terminology interprets these keywords and their parameters to the more arcane
-instruction set required by the underlying processor to perform some
-computation, such as invoking an executable machine code model. These shells
-are pre-compiled parsers, so were historically limited in terms of their
-expressibility and vocabulary size. Hence, they are generally less expressive
-than the virtual machines discussed above, but are still Turing complete. This
-means the coding of complex algorithms can be achieved but is awkward and slow
-because the keywords are not optimised for general purpose computation, but for
-carrying out operations at level of the operating system. For example, file
-control, with atomicity at the granularity of a single file can be done very
-efficiently using these kinds of languages. Because the latter is effectively
-utilising the operating system of the computer directly, then this grouping of
-languages is generally used to control an existing computer. That is, they are
-used as \emph{job control languages} (JCL). Examples of such languages are
-Bash, Ash, Dash, zsh, csh, Windows Power Shell, MSDOS, etc. Beyond the fact
-that scripting provides automation for large experiment sets, it also allows a
-certain degree of replicability, in that you to reproduce the experiment, then
-all that is required is that the scripts are rerun. However this is
-repeatability rather than replication. The aim is to just to rerun the
-experiment, rather than reproduce a given set of results.  We believe that
-anybody who does serious agent-based modelling experimentation should at the
+The automation of experiments is already solved by
+\textit{scripting}. A script -- in a scripting language such as Python, Julia, or
+Perl, but more typically an operating system `shell' language such as Bash --
+records the sequence of programs that need to be executed to reproduce the experiment.
+This at least repeats the
+experiment, rather than \textit{precisely} replicating a specific set of results.
+We believe that
+anybody who does experiments with their agent-based models should at the
 very least be scripting most of their experiments. In our experience the
-preparation and the execution of the experiment is relatively straight-forward
-to script. Post processing to the final results of data, diagrams, etc., tends
-to be a little neglected in the hurry to publish results.  Scripting is usually
-(although not exclusively) a text based activity, which means it is amenable to
-being programmatically produced.
+preparation and the execution of the experiment is relatively straightforward
+to script. Post-processing to the final data tables, diagrams, etc., that appear in
+manuscripts and other documents should be included in these scripts. These scripts
+can be conceived as \textit{workflow} metadata -- the means by which a \textit{class}
+of outputs is achieved, rather than a specific instance.\footnote{In the case that all of the programs executed are deterministic, the class is a singleton.}
 
-The second problem of recording of data, we decided on the use of a relational
-database. This was selected on the basis that such technologies have a proven
-track record of recording and manipulating reasonably large volumes of data.
-This is an established technology. It is also portable and many proven
-libraries exist for the manipulation and use of such databases. And it is this
-context that the discussion of scripting languages and virtual machines becomes
-relevant. The approach we utilised when manipulating this database was a
-scripting language that utilised a virtual machine, specifically Python. As
-mentioned this gives a portability allowing us to develop on laptops and run
-the resultant code in high-performance computing environments with little, or
-no reconfiguration.
+\textit{Provenance} metadata is used to record the means by which a specific output is achieved, which is more suitably stored in a 
+database. That database is updated using a series of scripts written in Python, allowing us to develop on laptops and run
+the resultant code in high-performance computing environments with little or
+no reconfiguration. Bash scripts automating the experiment workflow are then modified to
+call the Python scripts to update the provenance database. We implemented the database in Sqlite3 for local
+development and PostgreSQL in high-performance infrastructure. In this way, we have created a provenance
+tool that can record provenance during workflow execution.
 
-Thus the solution we selected was to use a proper JCL, in this case Bash to
-provide the necessary automation and replication models and call the provenance
-functions, written in a popular scripting language (Python) to update a
-standard relational database. In this case we selected Sqlite3 for local
-development and PostgreSQL for more heavy lifting in a distributed environment
-such as in high-performance infrastructure. Thus we have created a provenance
-tool which is based purely on scripting tools and a standard kind of database.
+\begin{figure} % wrapfigure is too small to see anything useful
+    \includegraphics[width=\textwidth]{img/schema.jpeg} % You should use a PNG at >= 300 dpi; not a lossy format like JPEG. You can export PNGs from PDFs using Preview
+\caption{SSREPI Schema} \label{fig:schema} \end{figure}
 
-\begin{wrapfigure}{l}{0.5\textwidth} 
-    \includegraphics[width=0.48\textwidth]{img/schema.jpeg}
-\caption{SSREPI Schema} \label{fig:schema} \end{wrapfigure}
+Another project adopting a similar approach is the \texttt{fair} data pipeline \cite{mitchell2022fair}, in which the provenance is automatically
+recorded by embedded model code in R, C++, Java, Julia and Python. However, one of our requirements
+was not to modify the model code if this could be avoided.\footnote{If the model code did not output the seed it used, then it would need to be modified.} In the example above, the FEARLUS-SPOMM model is written
+in Objective-C \cite{objectivec}; with additional supporting Perl and R scripts to prepare the data and conduct the
+post analysis. We would also prefer not to modify these scripts. Instead, we wrote wrapper scripts in Bash
+\cite{bash4420}. A further observation is that our framework
+generalises to more source metadata (papers, data, other experiments), and
+is written with the requirements of agent-based modelling in mind first and foremost.
 
-Since the inception of this project, we have subsequently uncovered a similar
-project which specifically concentrates on provenance data. This is the FAIR
-data pipeline \cite{mitchell2022fair} in which the provenance is automatically
-recorded by embedded code in R, C++, Java, Julia and Python.  One of the aims
-was not to modify the original code, i.e. the original experiment as unchanged
-as possible. That is we wanted to change the framework code rather than the
-core code. The core code in the example above being the executable, pre-compiled
-in Objective-C \cite{objectivec}. Additionally the support code, i.e. the code
-that prepared the data, a Perl \cite{perl563} script and the code that did the
-post analysis, a mixture of Perl and R \cite{R422}.  This meant we developed
-the database access in Python \cite{python368}, wrote wrapper scripts in Bash
-\cite{bash4420}. This differs in approach to [ibid] which is embedded in the
-code that constitutes the body of the experiment. Moreover our framework
-generalises to more more source metadata (paper, data, other experiments), and
-sis written with the requirements of agent-based modelling first and foremost.
-That is not to say that we do not have similar ambitions to automatically place
-provenance tracing code automatically embedded in experimental model code.
-
-The rest of this paper, we describe the scripting tool we have developed  for
+In the rest of this paper, we describe the scripting tool we have developed  for
 automatically recording metadata, which can be incorporated into the analysis
 replication process, and how this  was used to regenerate some of the figures
-in \cite{polhill_nonlinearities_2013}.
-In addition we show a simple tool we have already developed to trace data through the provenance graph, and as usual suggest additional work we would like to pursue and other ideas.
+in \cite{polhill_nonlinearities_2013}. In addition we show a simple tool we have already developed to trace data through the provenance graph, and suggest additional work we would like to pursue.
 
-\section{Method}
+\section{The \texttt{SSREPI} provenance and workflow tool}
 
 One of the artefacts of the MIRACLE project \cite{parker2019final} was the
-Social Simulation REplication Interface or SSREPI. This is the schema shown in
-figure \ref{fig:schema}.  This schema has evolved since it original conception.
-The newest version of this document may be found here
-\cite{polhill2022miracle}. This is a database schema derived from the Dublin
-Core \cite{weibel2000dublin}, the standard XML datatypes \cite{biron2004xml}
-and the PROV-O ontologies \cite{missier2013w3c}.  The schema is designed with
-agnosticisim towards the underlying database technology and as such has been
-implemented both in PostgresSQL \cite{stonebraker1991postgres} and Sqlite3
+Social Simulation REplication Interface or \texttt{SSREPI}. This is the schema shown in
+figure \ref{fig:schema}.  This schema has evolved since it original conception; 
+\cite{polhill2022miracle} is the latest version. The schema is derived from the Dublin
+Core \cite{weibel2000dublin} and PROV-O ontologies \cite{missier2013w3c} among others, and is designed with
+agnosticism towards the underlying database technology, having been
+implemented in PostgresSQL \cite{stonebraker1991postgres} and Sqlite3
 \cite{sqliteorg2023syntax}.
 
-There are two important dimensions of distinction to this schema. First is
-fine- versus coarse- grained metadata. Second is provenance versus workflow.
-Coarse-grained metadata describes how particular files come (or came) into
+Besides the distinction between workflow and provenance described above, we can discriminate
+fine- versus coarse- grained metadata. Coarse-grained metadata describes how particular files come (or came) into
 being, or were (or could be) used to bring other files into being. Fine-grained
@@
 data within the CSV file are covered by fine-grained metadata, whilst the fact
-that the simulation produces the CSV file is coarse-grained. Turning to the
-other dimension, provenance metadata describes what actually happens (run W of
-simulation X produced output file Y), whilst workflow metadata describes what
-could happen (simulation X produces an output file of type Z). The distinctions
-are summarised in \ref{fig:finegrain} .
+that the simulation produces the CSV file is coarse-grained.
 
-\begin{wrapfigure}{r}{0.5\textwidth}
-\includegraphics[width=0.5\textwidth]{img/fine-grain-vs-coarse-grain.jpeg}
-\caption{Fine grain vs coarse grain provenance} \label{fig:finegrain}
-\end{wrapfigure}
-
-Each table in schema shown in fig. \ref{fig:schema} was coded as an object type
+To build \texttt{SSREPI}, each table in schema shown in fig. \ref{fig:schema} was coded as an object type
 in Python. Each table row was represented as an instance of such an object.
 This design approach was adopted to enforce a consistent coding methodology
-across all tables.
+across all tables. To these were added a few simple coordinating commands that could be called from Bash:
 
-This then just left the interface between the two scripting languages. The
-design decision was taken to do domain specific processing in the Bash
-scripting language. This was done for due to the repetitive and time consuming
-nature of system testing: amendments in the job control language being easier
-to do quickly. This is a decision that may need revisiting. The problem being
-that Bash invocations are computationally and time expensive. Complex
-processing should be abstracted away from the slower language, in this case job
-control language to improve speed and remove what might be regarded as
-inappropriate computation for such an environment. Consequently as the system
-stands the commands that coordinated between Bash and Python are few and
-simple, the harder processing having been pushed down into the job control
-level, i.e. in the Bash scripts.
-
-These coordinating commands are:
-
-\begin{itemize} \item {\color{green} \ttvar{create_database.py}} - creates a
-        database idempotently.  \item {\color{green} \ttvar{exists.py}} - checks
-        if a particular row in a table exists.  \item {\color{green}
+\begin{itemize}
+    \item {\color{OliveGreen} \ttvar{create_database.py}} - creates a
+        database idempotently.
+    \item {\color{OliveGreen} \ttvar{exists.py}} - checks
+        if a particular row in a table exists.
+    \item {\color{OliveGreen}
             \ttvar{get_value.py}} - gets any specified single value from a
-            table given the primary key.  \item {\color{green}
-\ttvar{get_values.py}}  - gets one or more rows given the search values. \item {\color{green}\ttvar{next_study.py}} - gets the next available and unique study number. \item
-{\color{green} \ttvar{update.py}} - idempotently updates a particular row in a
+            table given the primary key.
+    \item {\color{OliveGreen} % 'green' hurt my eyes!
+\ttvar{get_values.py}}  - gets one or more rows given the search values.
+    \item {\color{OliveGreen}\ttvar{next_study.py}} - gets the next available and unique study number. \item
+{\color{OliveGreen} \ttvar{update.py}} - idempotently updates a particular row in a
 table.  \end{itemize}
 
-Note the use of idempotent, indicating that multiple operations on
+`Idempotent' indicates that multiple operations on
 the same entity will leave it unchanged after the initial operation. For
-example multiplying something by 1 is idempotent. This allows for a less
-rigorous exception criteria, for instance if a row or entity already exists.
-This implies that initialisation must be performed with care, as older data
-will not necessarily be destroyed but overwritten with newer values, but
-also retaining any values that already exist. Again this design decision was
-expedient on rapid development and may need revisiting. However the schema is
-and its interaction is quite complex and this would involve critical path
-analysis
+example multiplying something by 1 is idempotent. This allows for less
+rigorous exception criteria, but implies that initialisation must be performed with care, as older data
+will not necessarily be destroyed when overwritten with newer values, but
+retain any values that already exist.\footnote{This is a design decision that may need revisiting.}
 
-So each of the Bash commands listed in table \ref{tab:commands} composes over
-the above python commands to populate the SSREPI database in a consistent and
-logical manner.
-
-For the purposes of this short paper we are concentrating on a demonstration of
-recording provenance. Indeed we use the example, mentioned in the introduction
-of \cite{polhill2022miracle}, and modified the original Bash \cite{bash4420}
-scripts to include the what we denote as \textit{primitives}.  In order to
-record provenance we coded the following primitives from the SSREPI interface
-definition \cite{polhill2022miracle}. As noted above these are implemented in a
-mixture of Python and Bash.
-
-\tiny \begin{longtable}{|l|p{2cm}|p{7cm}|} \hline Primitive & Type & Purpose \\
-    \hline {\color{blue} \ttvar{SSREPI_require\_minimum}} & Metadata & Lower
+\begin{table}\begin{tabular}{|l|p{1cm}|p{5.5cm}|} \hline Primitive & Type & Purpose \\
+    \hline {\color{blue} \ttvar{SSREPI_require_minimum}} & M & Lower
     bound on software hardware required \\ {\color{blue}
-    \ttvar{SSREPI_require\_exact}} & Metadata & Exact bound on software
-    hardware required \\ {\color{blue} \ttvar{SSREPI_application}} & Provenance
-    \& Metadata & specifies some executable \\ {\color{blue} \ttvar{SSREPI_me}}
-    & Provenance \& Metadata & Determines executable being run or returns a
+    \ttvar{SSREPI_require_exact}} & M & Exact bound on software
+    hardware required \\ {\color{blue} \ttvar{SSREPI_application}} & P
+    \& M & specifies some executable \\ {\color{blue} \ttvar{SSREPI_me}}
+    & P \& M & Determines executable being run or returns a
     proper reference to the executable being run. \\ {\color{blue}
-    \ttvar{SSREPI_run}} & Provenance \& metadata & Blocking invocation of an
-    executable which will allow the specification of all inputs, outputs and
-    arguments. Creates run-time provenance information as well \\ {\color{blue}
-    \ttvar{SSREPI_batch}} & Provenance \& metadata & Non blocking invocation of
-    an executable which will allow the specification of all inputs, outputs and
-    arguments. Creates run-time provenance information as well \\ {\color{blue}
-    \ttvar{SSREPI_argument}} & Provenance & An argument type to an exectuable
-    \\ {\color{blue} \ttvar{SSREPI_output}} & Provenance & An output type from
-    an executable \\ {\color{blue} \ttvar{SSREPI_input}} & Provenance & An
-    input type for an exectuable \\ {\color{blue}
-    \ttvar{SSREPI_hutton\_person}} & Metadata & Uses our institutions databases
-    to populate metadata for a given individual \\ {\color{blue}
-    \ttvar{SSREPI_person}} & Metdata & Provide metadata for a particular actor
-    within this system\\ {\color{blue} \ttvar{SSREPI_project}} & Metadata &
+    \ttvar{SSREPI_argument}} & P & An argument type to an executable
+    \\ {\color{blue} \ttvar{SSREPI_output}} & P & An output type from
+    an executable \\ {\color{blue} \ttvar{SSREPI_input}} & P & An
+    input type for an executable \\ {\color{blue}
+    \ttvar{SSREPI_person}} & M & Provide metadata for a particular actor
+    within this system\\ {\color{blue} \ttvar{SSREPI_project}} & M &
     Specifies a project which contains all studies \\ {\color{blue}
-    \ttvar{SSREPI_study}} & Metadata & A set of experiments makes up a single
-    study \\ {\color{blue} \ttvar{SSREPI_set}} & Metdata & Sets the default
+    \ttvar{SSREPI_study}} & M & A set of experiments makes up a single
+    study \\ {\color{blue} \ttvar{SSREPI_set}} & M & Sets the default
     licence and other metadata \\ {\color{blue} \ttvar{SSREPI_involvement}} &
-    Metadata & Links personnel to a study \\ {\color{blue}
-    \ttvar{SSREPI_paper}} & Metadata & A paper associated with this study \\
-    {\color{blue} \ttvar{SSREPI_make_tag}} & Metadata & Used for building a
-    folksonomy \\ {\color{blue} \ttvar{SSREPI_tag}} & Metadata & Used to tag
+    M & Links personnel to a study \\ {\color{blue}
+    \ttvar{SSREPI_paper}} & M & A paper associated with this study \\
+    {\color{blue} \ttvar{SSREPI_make_tag}} & M & Used for building a
+    folksonomy \\ {\color{blue} \ttvar{SSREPI_tag}} & M & Used to tag
     any entity with a folksonomy tag \\ {\color{blue}
-    \ttvar{SSREPI_contributor}} & Metadata & A  person with some kind of
+    \ttvar{SSREPI_contributor}} & M & A  person with some kind of
     relation to an executable or script. \\
 
-    {\color{blue} \ttvar{SSREPI_statistical_method}} & Metadata & A statistical
-    method is an approach to computing some statistics. It may be implemented
-    in or as part of an application. A statistical method generates one or more
-    statistical variables as its results, and may use the results of another
-    statistical method in its computation. For example, computing the standard
-    deviation of some data uses the mean of those data. \\ {\color{blue}
-    \ttvar{SSREPI_visualisation}} & Metadata & A visualisation is the process
-    of creating an image to depict one or more (typically more than one)
-    values. The results of a visualisation appear in a container.\\
-    {\color{blue} \ttvar{SSREPI_statistics}} & Metadata & Statistics are
-    activities that compute  and populate the values of statistical rvariables.
-    They operate on raw data that are retrieved from the values using a query.
-    To replicate a set of statistics, the query can be rerun, selecting values
-    that are pointed to by containers entries \\ {\color{blue}
-    \ttvar{SSREPI_visualisation_method}} & Metadata & This describes methods
-    for generating visualisations, which then may appear in the content of a
-    container produced by a process running an application that implements it.
+    {\color{blue} \ttvar{SSREPI_statistical_method}} & M & Record a statistical method \\ {\color{blue}
+    \ttvar{SSREPI_visualisation}} & M & Record a method to create an image to depict one or more 
+    values. \\
+    {\color{blue} \ttvar{SSREPI_statistics}} & M & Record
+    activities that compute and populate the values of statistical variables. \\ {\color{blue}
+    \ttvar{SSREPI_visualisation_method}} & M & Methods
+    for generating visualisations.
     \\
 
-    {\color{blue} \ttvar{SSREPI_implements}} & Metadata & Links a statistical
+    {\color{blue} \ttvar{SSREPI_implements}} & M & Links a statistical
     or visualisation method to an application \\ {\color{blue}
-    \ttvar{SSREPI_parameter}} & Metadata & A Parameter is the name of a
-    parameter taken by a statistical or visualisation method, used to configure
-    the way it behaves. \\ {\color{blue} \ttvar{SSREPI_statistical_variable}} &
-    Metadata &  A name for (one of) the result(s) of a statistical method. \\
-    {\color{blue} \ttvar{SSREPI_visualisation_variable}} & Provenance \&
-    Metadata & Declares a named variable of interest \\ {\color{blue}
-    \ttvar{SSREPI_variable}} & Metadata & Names a variable of interest \\
-    {\color{blue} \ttvar{SSREPI_statistical_variable_value}} & Provenance \&
-    Metdata & Sets an actual value for a named statistical variable \\
-    {\color{blue} \ttvar{SSREPI_value}} & Provenance & Sets an actual value.
-    This can be connected to any metadata value such         \\ {\color{blue}
-    \ttvar{SSREPI_content}} & Metadata & Links a kind of output/input/argument
+    \ttvar{SSREPI_parameter}} & M & Record the name of a
+    parameter taken by a statistical or visualisation method. \\ {\color{blue} \ttvar{SSREPI_statistical_variable}} &
+    M &  A name for (one of) the result(s) of a statistical method. \\
+    {\color{blue} \ttvar{SSREPI_visualisation_variable}} & P \&
+    M & Declares a named variable of interest \\ {\color{blue}
+    \ttvar{SSREPI_variable}} & M & Names a variable of interest \\
+    {\color{blue} \ttvar{SSREPI_statistical_variable_value}} & P \&
+    M & Sets an actual value for a named statistical variable \\
+    {\color{blue} \ttvar{SSREPI_value}} & P & Sets a value.         \\ {\color{blue}
+    \ttvar{SSREPI_content}} & M & Links a kind of output/input/argument
     to a variable  \\ {\color{blue} \ttvar{SSREPI_person_makes_assumption}} &
-Metadata & links a person to an assumption \\ \hline \caption{Available Bash
-commands for provenance and metadata gathering} \label{tab:commands}
-\end{longtable}
+M & Links a person to an assumption \\ \hline \end{tabular} \caption{\texttt{SSREPI}'s Bash
+commands for provenance (Type column entry `P') and other metadata (Type `M') recording} \label{tab:commands}
+\end{table}
 
@@
 
-Broadly speaking {\color{blue} \ttvar{SSREPI_application}}, {\color{blue}
+Each of the Bash commands listed in table \ref{tab:commands} composes over % What does 'composes over' mean?
+the above Python commands to populate the \texttt{SSREPI} database in a consistent and
+logical manner. Broadly speaking, {\color{blue} \ttvar{SSREPI_application}}, {\color{blue}
 \ttvar{SSREPI_run}}, {\color{blue} \ttvar{SSREPI_batch}}, {\color{blue}
@@
 and {\color{blue} \ttvar{SSREPI_batch}} record fine-grain provenance. The
-remaining primitives are largely about recording metadata.
+remaining primitives are largely about recording other metadata. Further information may be found
+in the tool's public repository \footnote{\url{https://github.com/DougSalt/ABM-metadata}}.
 
-With the tool in the current state this involved laboriously going through the
-code and inserting the some of the Bash-based commands, shown in
-table \ref{tab:commands}. These commands have a fairly complex parameter set as
-it stands, and no way of checking whether the code worked other than running
-it, initially on a small subset of the experiment and then scaling up. This was
-extremely time-consuming.
+Using \texttt{SSREPI} for existing workflow scripts is a somewhat laborious task entailing adding calls to the provenance metadata scripts in Table \ref{tab:commands}.
+Improvements to the `interface' is the subject of future work.
 
-Elaboration of all this may be found in the man pages and design documents in
-the public repository at \url{https://github.com:/DougSalt/ABM-metadata}.
+\section{Demonstration}
 
-\begin{figure} \includegraphics[width=\textwidth]{img/workflow.pdf}
-\caption{The workflow sub-graph} \label{fig:workflow} \end{figure}
-
-We developed a very basic scheduler for use in an high performance computing
-environment, where there are many resources available both on the local
-machine. In particular for the situation where they is a single machine that
-has a large memory and many cores, the non-blocking version of the  two
-execution primitives, specifically {\color{blue} \ttvar{SSREPI_batch}}
-implements a kind of a simple scheduler locally using the sub-processing
-commands available to Unix-like systems. This was an historical solution to
-schedule in an high performance computing environment that existed prior to
-clustered computing situations. Clustering is the utilisation of one or more
-instances of such powerful computing machinery. Generally schedulers have been
-developed for such environments, as such SGE \cite{gentzsch2001sun} and the one
-we use as default, Slurm \cite{yoo2003slurm}. Scheduling is now largely handled
-outwith the framework, which makes the code less complex.
-
-{\color{blue} \ttvar{SSREPI_run}}, the other executing primitive will block
-until the command, code or model is has invoked has completed. This is
-necessarily used in single threaded coordination context, such as gathering all
-outputs necessary before proceeding to the next stage of processing. For
-instance, initial post processing of an all experiments would represent such a
-single threaded context.
-
-\section{Results}
-
-The primary result is that the target results and diagrams from the original
-paper \cite{polhill2017lessons} were able to be recreated.  As mentioned the
-method undertaken in this paper involved inserting the correct Bash-based
-commands, listed in table \ref{tab:commands}. These commands have a fairly
-complex parameter set as it stands, and no way of checking whether the code
-worked other than running it. This led to one of the immediate problems with
-this method. The only way to see if the code was working was to run it, a suck
-it and see methodology. This was extremely time consuming. We took the obvious
-approach and ran small subsets of the ~20,000 runs or so, to see if we could
-get the framework to work and then scaled it up to the full run, but invariably
-we ran into new problems the more we increased the scale. It therefore took
-substantial time to even get this first replication to run. It took several
-months of coding effort to get this correct. Generally it would alternate
-between parameter errors for the calls in table \ref{tab:commands}, and errors in the scripts making calls to record the provenance data to the database.
-Each iteration would take weeks to run, before failure as we increased the scale.
+Here, we demonstrate the use of \texttt{SSREPI} by repeating the experiment in the
+paper \cite{polhill2017lessons}, showing some of the workflow and provenance graphs, and then using fictitious examples to apply the workflow and provenance metadata to addressing issues in large-scale experiments.
+From a workflow perspective, the main result is that the
+principal diagrams in that paper were successfully recreated, with the same, albeit
+not identical, results. %right? % Plus, I've deleted all the whining :-)
 
 After a run had completed successfully the database was run through several
-programs to produce the sub-graphs detailed in fig. \ref{fig:schema}. These
-programs are capable of producing the following subgraphs:
+supplementary programs to produce the following visualisations of the provenance metadata recorded:
 
@@
 
-We developed programs that took the relational data in order to produce "dot"
-files. Such files that can be processed by Graphviz
-\cite{ellson2002graphviz}, and were used to produce the diagrams in figures
+The supplementary programs use the Dot language for input to Graphviz\cite{ellson2002graphviz}, producing the diagrams such as those in figures
 \ref{fig:sub-provenance}, \ref{fig:workflow}, \ref{fig:broken-application} and
-\ref{fig:high-lit-provenance-trace} .
+\ref{fig:high-lit-provenance-trace}. The workflow graph is in
+fig.
+\ref{fig:workflow}.
+The resultant provenance graph is massive, since there were 20,000 runs (so 20,000 sets of outputs to record), so we only show a \textit{very small} section of the provenance graph in fig.
+\ref{fig:sub-provenance}.
 
-To give an idea of the visualisation available we show the workflow graph in
-fig. \ref{fig:workflow}. This is a section of the workflow and is relative easy
-to follow. The resultant provenance graph is both massive and massively
-complex, given there were 20,000 runs (so 20,000 sets of outputs to record), so we only show a \textit{very small} section of the provenance graph in fig.
-\ref{fig:sub-provenance}.
+\begin{figure} % I suggest you put 'dir = LR' in this DOT file -- we can't really read the workflow
+\includegraphics[width=\textwidth]{img/workflow.pdf}
+\caption{The workflow sub-graph} \label{fig:workflow} \end{figure}
 
@@
 \includegraphics[width=\textwidth]{img/subsection-of-provenance.png}
-\caption{A small sub-section of the proevenance graph} \label{fig:sub-provenance}
+\caption{A small sub-section of the provenance graph} \label{fig:sub-provenance}
 \end{figure}
 
-So what use is this provenance metadata? Beyond replicability? Say for instance
-we have a bad dataset. The challenge is not only how to visualise the
-provenance, but to use this data in a useful manner. We propose a simple use
-here.  The "dot" language used by Graphviz constitutes a primitive graph
-database. Indeed there are programs that can transform "dot" files into
-TinkerPop GraphSON format \cite{tinkerpop2015graphson,gv2graphson}. Rather than
-using the relational database data we used the resultant dot files listed above
-to trace bad entities. \footnote{These are directed graphs (unproven as to
-whether they are acyclic - they shouldn't be they are the result of a
-relational database which was designed to be and \textit{should} be acyclic).}
-
-So for an easy to see example we pick the entity {\color{red}\ttvar{Applications.application_3831436655}}, and see how this affects other entities in the graph.
-This is \ttvar{SSS-StopC2-Cluster-create.pl} and we are using the workflow graph to see what might be affected.
-And this is shown in \ref{fig:broken-application}.
-It might be for instance that there is a bug in this program, so we should like to see how the errors might propagate out through the workflow.
+So what use is this provenance metadata? Imagine for the purposes of illustration that
+we have found a bug in a script.
+The entity {\color{red}\ttvar{Applications.application_3831436655}} is an experiment setup script \ttvar{SSS-StopC2-Cluster-create.pl}.
+We can use the workflow graph to see what else in the workflow might be affected.
+As might be expected for an experiment setup script, there are serious cascading consequences, which we can visualise in Figure \ref{fig:broken-application}.
+The Dot language used by Graphviz constitutes a primitive graph
+database. Indeed there are programs that can transform Dot files into
+TinkerPop GraphSON format \cite{tinkerpop2015graphson,gv2graphson}. To generate Figure \ref{fig:broken-application}, we used the workflow visualisation Dot file rather than the relational database.
 
 \begin{figure} \includegraphics[width=\textwidth]{img/broken-application.pdf}
-\caption{The workflow sub-graph show the propagation of a broken application (in red) to its outputs} \label{fig:broken-application} \end{figure}
+\caption{The workflow sub-graph show the propagation of a broken application (in red) to anything transitively making use of its outputs} \label{fig:broken-application} \end{figure}
 
-That example was somewhat contrived and simplistic and gives a flavour.
-However, what if we wanted to trace a bad data set?  If we ignore for this
-original example that the datasets for the above example are generated
-automatically by Perl scripts, and we pick one randomly to illustrate how this
-approach might work. The dataset in question would be:
-{\color{red}\ttvar{Containers.container_42949672955}}, a data file that forms
-a input file to just one experiment.  And here we see, again, a \textit{very
-small} section of of the provenance graph showing what is affected in red. This
-is shown in fig. \ref{fig:high-lit-provenance-trace}
+Similarly, the provenance metadata can be used to check the damage caused to a large-scale experiment by a single bad dataset.
+In the real experiment, the input files are generated by running a script, but for the purposes of the example, suppose the data in
+{\color{red}\ttvar{Containers.container_42949672955}} is wrong. The data in this file forms
+a input file to a handful of runs, and we want an idea of how it has affected our results.
+We can visualise the propagation of the error in a \textit{very small} section of of the provenance graph in fig.
+\ref{fig:high-lit-provenance-trace}, but we can use the database to list the entities affected:
+
+\tiny
+
+\begin{table}
+\begin{tabular}{ll}
+	\ttvar{Applications.application_3450918915} &
+	\ttvar{Applications.application_648609270} \\
+	\ttvar{Computers.asterix.local} &
+	\ttvar{Containers.container_1814970370} \\
+	\ttvar{Containers.container_1982026419} &
+	\ttvar{Containers.container_2050039078} \\
+	\ttvar{Containers.container_2056384913} &
+	\ttvar{Containers.container_2060874102} \\
+	\ttvar{Containers.container_2387213333} &
+	\ttvar{Containers.container_2486610989} \\
+	\ttvar{Containers.container_2582525701} &
+	\ttvar{Containers.container_2759060318} \\
+	\ttvar{Containers.container_2865400753} &
+	\ttvar{Containers.container_3025688835} \\
+	\ttvar{Containers.container_3307537171} &
+	\ttvar{Containers.container_3470971297} \\
+	\ttvar{Containers.container_354343442} &
+	\ttvar{Containers.container_4235735972} \\
+	\ttvar{Containers.container_4294967295} &
+	\ttvar{Containers.container_441913555} \\
+	\ttvar{Containers.container_505627104} &
+	\ttvar{Containers.container_800277554} \\
+	\ttvar{Containers.container_878886043} &
+	\ttvar{Containers.container_900718909} \\
+	\ttvar{Persons.doug} &
+	\ttvar{Persons.doug_salt} \\
+	\ttvar{Processes.process_232221298886493}... &
+	\ttvar{Processes.process_326475499597022}... \\
+	\ttvar{Specifications.R} &
+	\ttvar{Specifications.bash} \\
+	\ttvar{Specifications.cpus} &
+	\ttvar{Specifications.disk_space} \\
+	\ttvar{Specifications.memory} &
+	\ttvar{Specifications.os} \\
+	\ttvar{Specifications.perl} &
+	\ttvar{Specifications.python} \\
+	\ttvar{Studies.1} &
+	\ttvar{Users.doug}
+ \end{tabular}
+\end{table}
+
+\normalsize
+
+Having done so, we are then in a position to fix the data file, and rerun only that part of the workflow needed to regenerate the affected containers.
 
@@
 
-Moreover we can list out the affected entities. In this case, those entities would be:
+\section{Future work}
 
-\tiny
-
-\begin{itemize}
-	\item \ttvar{Applications.application_3450918915}
-	\item \ttvar{Applications.application_648609270}
-	\item \ttvar{Computers.asterix.local}
-	\item \ttvar{Containers.container_1814970370}
-	\item \ttvar{Containers.container_1982026419}
-	\item \ttvar{Containers.container_2050039078}
-	\item \ttvar{Containers.container_2056384913}
-	\item \ttvar{Containers.container_2060874102}
-	\item \ttvar{Containers.container_2387213333}
-	\item \ttvar{Containers.container_2486610989}
-	\item \ttvar{Containers.container_2582525701}
-	\item \ttvar{Containers.container_2759060318}
-	\item \ttvar{Containers.container_2865400753}
-	\item \ttvar{Containers.container_3025688835}
-	\item \ttvar{Containers.container_3307537171}
-	\item \ttvar{Containers.container_3470971297}
-	\item \ttvar{Containers.container_354343442}
-	\item \ttvar{Containers.container_4235735972}
-	\item \ttvar{Containers.container_4294967295}
-	\item \ttvar{Containers.container_441913555}
-	\item \ttvar{Containers.container_505627104}
-	\item \ttvar{Containers.container_800277554}
-	\item \ttvar{Containers.container_878886043}
-	\item \ttvar{Containers.container_900718909}
-	\item \ttvar{Persons.doug}
-	\item \ttvar{Persons.doug_salt}
-	\item \ttvar{Processes.process_23222129888649347509041510897074}
-	\item \ttvar{Processes.process_32647549959702293885495865014666}
-	\item \ttvar{Specifications.R}
-	\item \ttvar{Specifications.bash}
-	\item \ttvar{Specifications.cpus}
-	\item \ttvar{Specifications.disk_space}
-	\item \ttvar{Specifications.memory}
-	\item \ttvar{Specifications.os}
-	\item \ttvar{Specifications.perl}
-	\item \ttvar{Specifications.python}
-	\item \ttvar{Studies.1}
-	\item \ttvar{Users.doug}
-\end{itemize}
-
-\normalsize
-
-\section{Discussion}
-
-What we have learnt from this?  As noted the process of running the experiment
-with the new provenance and metadata commands was time-consuming to the point
-of almost being unusable. It involved the repeated running of the experiment
-with a larger and larger array of inputs until we matched the size of the
-original experiment.  This has taken several months of intense work to event get it running. Repeating an experiment in a completely automated fashion turns out to be a really hard task.
-
-However, we have to emphasise that this was the reproduction of the experiment.
-If this framework were available from the start then most of the work would be
-done once the experiments had been completed; that is, the work needed to
-actually repeat the experiment in a consistent and reliable manner. This is the
-key, and it reduces to the same advice that meta-data advocates everywhere
-proclaim, which is to incorporate the metadata from the start and not as an
-after thought.
-
-Provenance should be a directed acyclic graph. There may well be loops in the
-schema as it stands. Given sufficient time we can certainly normalise the
-relational database to remove redundancy and the conflict arising from such,
-and once this complete,  we can formally prove the schema to be acyclic.  The
-authors have done their best to normalise it to uncover such cycles, but have
-yet to formally do so. In a schema this complex they may well still exist.
+Provenance should be a directed acyclic graph. As the schema stands, it does not
+guarantee that the graph is acyclic. In future work, we can  normalise the
+schema to remove redundancy and the conflicts arising therefrom;
+and once this complete, we can formally prove the schema to be acyclic. 
 Until the schema is formally analysed, then we can not say with confidence that
-the schema is actually consistent, in the sense that any recorded provenance is
-not inherently contradictory. Given that such a schema is likely to be
-continually revised as the underlying provenance model changes or additional
+any recorded provenance is not inherently contradictory. Since such a schema is likely to be
+iteratively revised if the underlying provenance model changes or additional
 functionality is required, then it would be advisable to automate such proving
@@
 
-In addition to reproducing the experiment automatically, there was some scope
-in utilising the parallelising infrastructure of the high performance computing
-environment.  One of the nice things was that we were able to parallelise part
-of the post processing job, and thus speed up the whole experiment as this was
-proving to be a bit of a bottleneck. However, careful thought needs to be given
-to this. In the case of agent-based modelling such stochasticism in inherent to
-the testing methodology, and it would be expected that such changes should have
-little impact. For other testing frameworks, particular those that do not rely
-on massive replication, then such changes may well have unintended
-consequences.  Indeed, those with a sharp eye may have noticed the versions of
-Perl, Python and R we use were not around when the original experiments were
-run and obviously this has implications. However, the framework does offer the
-ability to record and enforce metadata about programming versions.
-
-The underlying database is almost certainly unsuitable. We should probably be
-using a graphing language such as Gremlin \cite{rodriguez2015gremlin}. The
+We should probably be using a graphing language such as Gremlin \cite{rodriguez2015gremlin}. The
 advantages of using graphing databases over relational databases is that such
@@
 trivial and fast. In a relational databases the SQL statement to do this are
-cumbersome, awkward and probably error prone given their size. Moreover on huge
+cumbersome, awkward and probably (human) error-prone given their size and complexity to construct. Moreover on huge
 datasets they are reportedly slow. Relational databases do have the advantage
 of being an extremely mature technology and the availability of utilities that
-implies. As mentioned, the advantage of using the relational database is that
-structured query language is standard for all such databases, and therefore
+implies. A further advantage of using relational databases is that
+Structured Query Language is standard for all such databases, and therefore
 queries written for any relational database should be the same. A tool we
-already use, Graphviz already does act like a graphing database language to a
-certain extent. Indeed we use the Graphviz "dot" files, rather than the
-relational schema in our simple tools tools to trace bad data through our
-provenance graphs.  However the "dot" language of Graphviz is primarily
+already use, Graphviz, already does act like a graphing database language to a
+certain extent. Indeed we used the Graphviz Dot files produced by the visualisation supplementary programs, rather than the
+relational schema, in our simple examples to trace bad data through our
+provenance graphs.  However the Dot language of Graphviz is primarily
 purposed as a diagramming language and lacks the sophistication, such as a
@@
 
-For future we work, we should eventually like to take many provenance databases run some machine learning across them to see if there are any commonalities.
+We would eventually like to take many provenance databases run some machine learning across them to see if there are any commonalities.
 We hope to uncover similarities in setup, execution and post-processing that could form core, reusable and proven components for primarily agent-based modelling but also for other modelling frameworks.
+These can then be used to \textit{suggest} workflow -- e.g.
+given files of these kinds, what have others done to visualise them?
 
@@
 frameworks, in particular R, Python, Julia, Java and thence NetLogo. This would
-take the approach of the FAIR data pipeline \cite{mitchell2022fair}, but unlike
+take the approach of the \texttt{fair} data pipeline \cite{mitchell2022fair}, but unlike
 the example we have presented here such provenance would be embedded in the
-experiments as a matter of course. We were constrained by the nature of the
-replication here to use Bash as our means of recording provenance. However this
-does show that such a provenance framework may be "retrofitted" to existing
-experiments (although possibly at the cost of the programmer's sanity given the
-minutiae involved). We have tentative potential adopters using R and Python as their primary modelling language.
+experiments as a matter of course. In the meantime, our demonstration with FEARLUS-SPOMM shows that the \texttt{SSREPI} provenance framework may be `retrofitted' to existing
+experiments (although possibly at the cost of the programmer's sanity with the present interface), and we can successfully repeat the published experiment.
+In the short-term, we have tentative potential adopters using R and Python as their primary modelling language.
 

### 2023-07-20 paper/paper.tex
Pre initial submission version
Updated paper for ESSA2023 reviewer comments

@@
 %
-% \author{Doug Salt\inst{1}\orcidID{0000-0001-5186-9388} \and
-% Gary Polhill\inst{1}\orcidID{0000-0002-8596-0590} }
+\author{Doug Salt\inst{1} \orcidID{0000-0001-5186-9388} \and
+ Gary Polhill\inst{1} \orcidID{0000-0002-8596-0590} \and
+ Corran Musk\inst{1} \and
+ Lorenzo Milazzo\inst{2} \and
+ Dawn Parker\inst{3} \orcidID{0000-0001-8988-193X}  }
 
@@
 %
-%\authorrunning{D. Salt et al.}
+\authorrunning{D. Salt et al.}
 % First names are abbreviated in the running head.
@@
 %
-% \institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
-% \email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
-%\url{https://www.hutton.ac.uk/}
-%}
+\institute{The James Hutton Institute, Craigiebuckler, Aberdeen, AB15 8QH, Scotland
+\email{\{doug.salt,gary.polhill\}@hutton.ac.uk}\\
+\url{https://www.hutton.ac.uk/}
+\and Satellite and Surface Assimilation (SSA) Group
+Met Office, FitzRoy Road, Exeter, Devon, EX1 3PB, UK
+\email{lorenzo.milazzo@physics.org}
+\and University of Waterloo, 200 University Avenue West
+Waterloo, ON, Canada  N2L 3G1
+\url{https://uwaterloo.ca/}
+}
 %
@@
 \cite{ludascher2006scientific} and Taverna \cite{hull2006taverna}. However, since
-these are workflow tools, the focus is on automation and
-repeatability rather than provenance, which, if it is included, is as an afterthought.
+these are workflow tools, the focus seems to be  on automation and
+repeatability rather than provenance, which, if it is included, seems not to be the primary purpose of the described tools.
+There are exceptions to this, such as the open provenance framework \cite{moreau2008open}, implemented in \cite{gadelha2011provenance} with yet another workflow control langauge \cite{zhao2007swift}, but such tools do not seem to have flourished and seemingly for agent-based models are non existent.
 
@@
 remaining primitives are largely about recording other metadata. Further information may be found
-in the tool's public repository \footnote{\url{https://github.com/DougSalt/ABM-metadata}}.
+in the tool's public repository. %\footnote{\url{https://github.com/DougSalt/ABM-metadata}}.
 
@@
 
-Here, we demonstrate the use of \texttt{SSREPI} by repeating the experiment in the
-paper \cite{polhill2017lessons}, showing some of the workflow and provenance graphs, and then using fictitious examples to apply the workflow and provenance metadata to addressing issues in large-scale experiments.
-From a workflow perspective, the main result is that the
-principal diagrams in that paper were successfully recreated, with the same, albeit
-not identical, results. %right? % Plus, I've deleted all the whining :-)
+Here, we demonstrate the use of \texttt{SSREPI} by repeating the experiment in
+the paper \cite{polhill2017lessons}, showing some of the workflow and
+provenance graphs, and then using fictitious examples to apply the workflow and
+provenance metadata to addressing issues in large-scale experiments. From a
+workflow perspective, the main result is that the principal diagrams in that
+paper were successfully recreated, with the same, albeit not identical,
+results. %right? % Plus, I've deleted all the whining :-)
+% whoops the reviewer complained about there being nothing here, so I am going
+% to have to reintrouduce some of the whining. :)
+By this we mean that the diagrams produced for the original paper
+\cite{polhill2017lessons} were "roughly the same". They were not identical, but
+contained similar features close enough to illustrate the features that were originally reported on in the original research from
+\cite{polhill2017lessons}.
 
@@
         \item \textbf{Finegrain} - a provenance diagram down to the level of
-        variables.  \item \textbf{Folksonomomy} - a diagram showing annotations
+        variables.  \item \textbf{Folksonomy} - a diagram showing annotations
             against the database, produced and categorised at the discretion of
@@
 The supplementary programs use the Dot language for input to Graphviz\cite{ellson2002graphviz}, producing the diagrams such as those in figures
-\ref{fig:sub-provenance}, \ref{fig:workflow}, \ref{fig:broken-application} and
+\ref{fig:workflow}, \ref{fig:sub-provenance}, \ref{fig:broken-application} and
 \ref{fig:high-lit-provenance-trace}. The workflow graph is in
@@
 The Dot language used by Graphviz constitutes a primitive graph
-database. Indeed there are programs that can transform Dot files into
-TinkerPop GraphSON format \cite{tinkerpop2015graphson,gv2graphson}. To generate Figure \ref{fig:broken-application}, we used the workflow visualisation Dot file rather than the relational database.
+database. Indeed there are programs that can transform Dot files into TinkerPop
+GraphSON format \cite{tinkerpop2015graphson,gv2graphson}, two reasonably well known and utilised graphing database formats. To generate Figure
+\ref{fig:broken-application}, we used the workflow visualisation Dot file
+rather than the relational database.
 
@@
 
+\section{Discussion}
+
+It might seem that we are confusing random stream control and reproducibility
+of individual simulation runs, i.e., given a set of inputs and random stream
+that the simulation would always produce the same results, versus the
+reproducibility of an overall analysis that comes out of a calibration or other
+simulation workflow. However, the latter issues cannot even begin to be
+addressed until a reproducible and reliable workflow is implmented for agent based models -
+that is a directed acyclic graph from inputs to published outputs/visualisation
+and/or results. This may seem like a contradiction to our earlier claim that
+this is not a workflow tool, but workflow in agent-based models is complex and
+the link from data/model to published material for agent-based models is
+currently missing. Agent based modelling does not have any specialist needs in
+terms of work control other than magnitude on the number of inputs, number of
+runs and number of resuts. However step ups in magnitude are not always trivial
+as the step up petaflop to exaflop is demonstrates
+\cite{north2008agent}.
+
 \section{Future work}

### 2023-07-20 paper/paper.tex
Changed diagrams and such like for the reviewers

@@
 \begin{figure} % wrapfigure is too small to see anything useful
-    \includegraphics[width=\textwidth]{img/schema.jpeg} % You should use a PNG at >= 300 dpi; not a lossy format like JPEG. You can export PNGs from PDFs using Preview
-\caption{SSREPI Schema} \label{fig:schema} \end{figure}
+    \includegraphics[width=\textwidth]{img/schema.png} % You should use a PNG at >= 300 dpi; not a lossy format like JPEG. You can export PNGs from PDFs using Preview
+    \caption{The overall SSREPI Schema with the workflow part magnified (will be used later in the paper}
+    \label{fig:schema} 
+\end{figure}
 
@@
 \begin{figure} % I suggest you put 'dir = LR' in this DOT file -- we can't really read the workflow
-\includegraphics[width=\textwidth]{img/workflow.pdf}
-\caption{The workflow sub-graph} \label{fig:workflow} \end{figure}
+\includegraphics[width=\textwidth]{img/workflow.png}
+\caption{The workflow sub-graph - showing the part of the graph of interest magnified} \label{fig:workflow} \end{figure}
 
@@
 \includegraphics[width=\textwidth]{img/subsection-of-provenance.png}
-\caption{A small sub-section of the provenance graph} \label{fig:sub-provenance}
+\caption{A small sub-section of the provenance graph - this is deliberately selected to indicate the complexity of the graph and how hard it is to read.} \label{fig:sub-provenance}
 \end{figure}
 
-So what use is this provenance metadata? Imagine for the purposes of illustration that
+So what use is this provenance metadata, if as shown in fig. \ref{fig:sub-provenance} it is so difficult to read or visualise? Imagine for the purposes of illustration that
 we have found a bug in a script.
@@
 \ref{fig:broken-application}, we used the workflow visualisation Dot file
-rather than the relational database.
+rather than the relational database. In addition we used the workflow sub-graph because this is probably one of the easier graphs to follow visually.
 
-\begin{figure} \includegraphics[width=\textwidth]{img/broken-application.pdf}
-\caption{The workflow sub-graph show the propagation of a broken application (in red) to anything transitively making use of its outputs} \label{fig:broken-application} \end{figure}
+\begin{figure} 
+    \includegraphics[width=\textwidth]{img/broken-application.png}
+    \caption{The workflow sub-graph show the propagation of a broken application (in red) in the background. The break-out box shows a magnified detail to show the content of the graph.} 
+    \label{fig:broken-application} 
+\end{figure}
 

### 2023-09-04 paper/paper.tex
Added expernses and finished presentation

@@
 \begin{abstract}
-    We demonstrate a working framework for the automatic recording of
+\    We demonstrate a working framework for the automatic recording of
     provenance and metadata for primarily agent-based models that could easily
@@
 
-\keywords{Provenance \and metadata \and modelling \and automation \and replication.}
+keywords{Provenance \and metadata \and modelling \and automation \and replication.}
 \end{abstract}
@@
 \begin{figure} \includegraphics[width=\textwidth]{img/high-lit-provenance-trace.pdf}
-\caption{A provenance sub-graph showing (in red) the propagation of bad data through the system} \label{fig:high-lit-provenance-trace} \end{figure}
+\caption{A provenance sub-graph showing (in red) the propagation of bad data through the system.
+Again we are being deliberately vague in order to give you an idea of the size and complexity of the graph} \label{fig:high-lit-provenance-trace} \end{figure}
 

### 2023-09-05 presentation/presentation.tex
Added the presentation

\documentclass[usenames,dvipsnames,10pt]{beamer}
%\documentclass[usenames,dvipsnames,10pt,handout]{beamer}

\usepackage{graphicx}
\usepackage{tikz}
\usepackage{longtable}
%\usepackage[usenames,dvipsnames]{xcolor}

\newcommand*\ttvar[1]{\texttt{\expandafter\dottvar\detokenize{#1}\relax}}
\newcommand*\dottvar[1]{\ifx\relax#1\else
  \expandafter\ifx\string_#1\string_\allowbreak\else#1\fi
  \expandafter\dottvar\fi}
%
%\usepackage{pgfpages}
\mode<handout>{%
    \pgfpagesuselayout{4 on 1}[a4paper]
    \setbeameroption{show notes}
}

\usepackage{hutton/hutton}

\title{Towards automated provenance collection for experimental runs of agent-based models}
\author{Doug Salt \and Gary Polhill\ and Corran Musk \and Lorenzo Milazzo \and Dawn Parker}
\institute{The James Hutton Institute}

\begin{document}

\begin{frame}[plain]
    \maketitle
\end{frame}

\begin{frame}
    \frametitle{Structure of the presentation}

    \begin{columns}
        \column{0.5\textwidth}
            \begin{itemize}
                \item Abstract
                \item Motivation
                \item The standard approach
                \item The database
                \item The scope
                \item A simple workflow
            \end{itemize}
        \column{0.5\textwidth}
            \begin{itemize}
                \item Some provenance
                \item So what?
                \item But so what?
                \item Maybe a use?
                \item What else?
            \end{itemize}
    \end{columns}
\end{frame}

\begin{frame}
    \frametitle{Abstract}
    We demonstrate a working framework for the automatic recording of
    provenance and metadata for primarily agent-based models that could easily
    be adapted to the other modelling environments. We discuss the need for
    such a framework, the philosophy behind the design we adopted, the
    implementation, discuss the results and demonstrate a simple tool for for
    tracing bad data through a provenance graph.
\end{frame}

\begin{frame}
    \frametitle{Motivation}

    \note{When I was a callow post-doc, my boss gave me a challenge - replicate this work}

    \begin{center}
        \begin{tikzpicture}
            \node (img1) {\includegraphics[height=3.5cm]{img/page3.png}};
            \node (img2) at (img1.north east) [yshift=1cm]{\includegraphics[height=3.5cm]{img/page2.png}};
            \node (img3) at (img2.south east) [yshift=1cm]{\includegraphics[height=3.5cm]{img/page1.png}};
        \end{tikzpicture}
    \end{center}
\end{frame}

\begin{frame}
    \frametitle{The standard approach}
    \huge
    \begin{itemize}
        \item Scripts 
        \item Storage archives
        \item Diaries
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{How we did it}
    \includegraphics[width=\columnwidth]{img/how.png}
\end{frame}

\begin{frame}
    \center
    \frametitle{The database}

    \includegraphics[width=.75\textwidth]{img/schema.png}

    \large
    \textbf{S}ocial \textbf{S}imultion \textbf{RE}plication \textbf{I}nterface
\end{frame}

\begin{frame}
    \frametitle{The scope}
    \begin{itemize} 
        \item \textbf{Analysis} - fine grain provenance pertaining to statistical and visualisation outputs.
        \item \textbf{Finegrain} - a provenance diagram down to the level of variables.  
        \item \textbf{Folksonomy} - a diagram showing annotations against the database, produced and categorised at the discretion of the user doing the annotation 
        \item \textbf{Project} - management metadata. Largest granularity of metadata supported 
        \item \textbf{Provenance} - provenance diagram at the level of file and parameter 
        \item \textbf{Services} - service provided and requirement description 
        \item \textbf{Workflow} - the actual workflow
    \end{itemize}
\end{frame}

\begin{frame}

    \frametitle{What was needed}
    \tiny
    \begin{table}
        \begin{tabular}{|l|p{1cm}|p{4.8cm}|} 
            \hline 
            Primitive & Type & Purpose \\
            \hline 
            {\color{blue} \ttvar{SSREPI_require_minimum}} & M & Lower bound on software hardware required \\ 
            {\color{blue} \ttvar{SSREPI_require_exact}} & M & Exact bound on software hardware required \\ 
            {\color{blue} \ttvar{SSREPI_application}} & P \& M & specifies some executable \\ 
            {\color{blue} \ttvar{SSREPI_me}} & P \& M & Determines executable or reference to executable being run. \\ 
            {\color{blue} \ttvar{SSREPI_argument}} & P & An argument type to an executable \\ 
            {\color{blue} \ttvar{SSREPI_output}} & P & An output type from an executable \\ 
            {\color{blue} \ttvar{SSREPI_input}} & P & An input type for an executable \\ 
            {\color{blue} \ttvar{SSREPI_person}} & M & Provide metadata for a particular actor within this system \\ 
            {\color{blue} \ttvar{SSREPI_project}} & M & Specifies a project which contains all studies \\
            {\color{blue} \ttvar{SSREPI_study}} & M & A set of experiments makes up a single study \\
            {\color{blue} \ttvar{SSREPI_set}} & M & Sets the default licence and other metadata \\
            {\color{blue} \ttvar{SSREPI_involvement}} & M & Links personnel to a study \\
            {\color{blue} \ttvar{SSREPI_paper}} & M & A paper associated with this study \\
            {\color{blue} \ttvar{SSREPI_make_tag}} & M & Used for building a folksonomy \\
            {\color{blue} \ttvar{SSREPI_tag}} & M & Used to tag any entity with a folksonomy tag \\
            {\color{blue} \ttvar{SSREPI_contributor}} & M & A  person with some kind of relation to an executable or script. \\
            {\color{blue} \ttvar{SSREPI_statistical_method}} & M & Record a statistical method \\
            {\color{blue} \ttvar{SSREPI_visualisation}} & M & Record a method to create an image to depict one or more values. \\
            {\color{blue} \ttvar{SSREPI_statistics}} & M & Record activities that populate the values of statistical variables. \\
            {\color{blue} \ttvar{SSREPI_visualisation_method}} & M & Methods for generating visualisations.  \\
            {\color{blue} \ttvar{SSREPI_implements}} & M & Links a statistical or visualisation method to an application \\
            {\color{blue} \ttvar{SSREPI_parameter}} & M & Record the name taken by a statistical or visualisation method. \\
            {\color{blue} \ttvar{SSREPI_statistical_variable}} & M &  A name for (one of) the result(s) of a statistical method. \\
            {\color{blue} \ttvar{SSREPI_visualisation_variable}} & P \& M & Declares a named variable of interest \\
            {\color{blue} \ttvar{SSREPI_variable}} & M & Names a variable of interest \\
            {\color{blue} \ttvar{SSREPI_statistical_variable_value}} & P \& M & Sets an actual value for a named statistical variable \\
            {\color{blue} \ttvar{SSREPI_value}} & P & Sets a value.         \\
            {\color{blue} \ttvar{SSREPI_content}} & M & Links a kind of output/input/argument to a variable  \\
            {\color{blue} \ttvar{SSREPI_person_makes_assumption}} & M & Links a person to an assumption \\
		 \hline 
        \end{tabular} 
    \end{table}
\end{frame}

\begin{frame}
    \frametitle{What was needed? (cont.)}
    We encoded the previous actions in Python and then then called these from Bash
    \begin{itemize}
        \item {\color{OliveGreen} \ttvar{create_database.py}} - creates a database idempotently.
        \item {\color{OliveGreen} \ttvar{exists.py}} - checks if a particular row in a table exists.
        \item {\color{OliveGreen} \ttvar{get_value.py}} - gets any specified single value from a table given the primary key.
        \item {\color{OliveGreen} \ttvar{get_values.py}}  - gets one or more rows given the search values.
        \item {\color{OliveGreen}\ttvar{next_study.py}} - gets the next available and unique study number. 
        \item {\color{OliveGreen} \ttvar{update.py}} - idempotently updates a particular row in a table.  
    \end{itemize}

\end{frame}

\begin{frame}
    \includegraphics[width=\textwidth]{img/workflow.png}
\end{frame}

\begin{frame}
    \frametitle{Some provenance!!!}
    \includegraphics[width=\textwidth]{img/subsection-of-provenance.png}
\end{frame}

\begin{frame}
    \frametitle{So what?}

    It now takes me the following lines of code to reproduce an experiment

    \scriptsize

    \begin{enumerate}
        \item Create new directory
        \item Enter new directory
        \item {\color{blue}\texttt{mkdir bin}}
        \item {\color{blue}\texttt{cp /mnt/storage/doug/git/ABM-metadata/example/*.sh bin}}
        \item {\color{blue}\texttt{cp /mnt/storage/doug/git/ABM-metadata/example/*.pl bin}}
        \item {\color{blue}\texttt{cp /mnt/storage/doug/git/ABM-metadata/example/*.R bin}}
        \item {\color{blue}\texttt{ln -s /mnt/storage/doug/git/ABM-metadata/cfg}}
        \item {\color{blue}\texttt{ln -s /mnt/storage/doug/git/ABM-metadata/lib}}
        \item {\color{blue}\texttt{ln -s /mnt/storage/doug/git/ABM-metadata/doc}}
        \item Amend {\color{blue}\texttt{bin/path.sh}} to {\color{blue}\texttt{bin/path.}}\textit{directory\_name}{\color{blue}\texttt{.sh}}
        \item source {\color{blue}\texttt{bin/path.}}\textit{directory\_name}{\color{blue}\texttt{.sh}}
        \item {\color{blue}\texttt{psql}}
        \item {\color{blue}\texttt{create database third}}
        \item {\color{blue}\texttt{\textbackslash{q}}}
        \item profit!
    \end{enumerate}

\end{frame}

\begin{frame}
    \frametitle{But so what?}
    \includegraphics[width=.8\textwidth]{img/broken-application.png}
\end{frame}

\begin{frame}
    \frametitle{Finally a use?}
    \includegraphics[width=\textwidth]{img/high-lit-provenance-trace.pdf}
\end{frame}

\begin{frame}
    \huge
    \center
    \huge
    What else?
    \note{
        1. We are thinking about running AI to look for patterns, if we have a sufficient number of provenace graphs
        2. Modularising and incorporating it into tools so this is automatically generated
        3. Better visualisation
        }
\end{frame}

\begin{frame}[plain]
    Thank you very much
    \finalpage
\end{frame}

\end{document}
